---
title: "SA256 DTG uptake <br> tblLAB_RNA, tblLAB_CD4"
subtitle: " "
author: ""
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    code_folding: show
    self_contained: true
    highlight: pygments
    keep_md: no
    toc_depth: 3
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
---

```{r r-setup, include = FALSE}
options(scipen = 999)
options(max.print = "75")
set.seed(12345)

library(pacman)
p_load(
  kableExtra,
  scales, ggplot2, dplyr,
  fst, data.table, sjmisc,
  osfr
)

import::from("psych", "geometric.mean")
import::from("sjmisc", "frq")
import::from("gmodels", "CrossTable")
```

```{r conflicts, include = FALSE}
conflicted::conflict_scout()
```

```{r knit-setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

knitr::opts_chunk$set(
  cache = FALSE,
  prompt = FALSE,
  tidy = FALSE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  echo = TRUE
)

knitr::opts_knit$set(width = 75)
```

<!-- ----------------------------------------------------- -->


```{r}
#Clear existing data and graphics
rm(list=ls())
graphics.off()
```



# IeDEA-SA Data Prep

Raw Stata and `fst` files come from central directory of `IeDEA_core`.  

Using release `Stata_202308_Aug`.  

```{r eval=FALSE, include=FALSE}
list.files("../IeDEA_core/data/Stata_202308_Aug/", pattern = "\\.fst$") %>%
  kable(col.names = "Datasets - new") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = F
  )
```

## Druglist

Updated `druglist` dataset from the `ART-codes` repo.   

```{r include=FALSE}
druglist = data.table(readr::read_csv("https://raw.githubusercontent.com/IeDEA-SA/ART-codes/main/data-raw/IeDEA_druglist.csv"))[, c("drug_orig") := NULL]

druglist[, drug := trimws(drug, which = "both")]
```

```{r echo=FALSE}
druglist %>%
  kbl() %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  column_spec(1, italic = TRUE) %>%
  row_spec(1, italic = FALSE)
```

```{r include=FALSE}
druglist = druglist[, .(art_id, drug, arv_class)]
```

## Official DTG adoption dates

Pulled by @aezaniewski from different sources.  

```{r include=FALSE}
dtg_recomend = data.table(read_fst("data_temp/dtg_recomend_v001.fst"))
```

```{r echo=FALSE}
dtg_recomend %>%
  kbl() %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  column_spec(1, italic = TRUE) %>%
  row_spec(1, italic = FALSE)
```


```{r}

rev_close_d = data.table(read_fst("data_temp/rev_close_d_v001.fst"))

program_close_d = data.table(read_fst("data_temp/program_close_d_v001.fst"))

tblCENTER = data.table(read_fst("data_temp/tblCENTER_v001.fst"))

tblBAS = data.table(read_fst("data_temp/tblBAS_v002.fst"))

tblLTFU = data.table(read_fst("data_temp/tblLTFU_v002.fst"))

tblART = data.table(read_fst("data_temp/tblART_v002.fst"))

```



<!-- ----------------------------------------------------- -->

# `tblLAB_RNA`

```{r}
tblLAB_RNA = data.table(read_fst("../IeDEA_core/data/Stata_202308_Aug/tblLAB_RNA.fst"))[, c("rna_t") := NULL]
```

Starting point is `r number(nrow(tblLAB_RNA), big.mark = ",")` ART records of `r number(length(unique(tblLAB_RNA$patient)), big.mark = ",")` patients from all combined cohorts.  

## BAS overlap

```{r}
tblLAB_RNA = merge(tblBAS[, .(patient, program)],
                   tblLAB_RNA,
                   by = c("patient"),
                   all.x = FALSE, all.y = FALSE
)
```

## Keep unique rows

```{r}
if (nrow(tblLAB_RNA) - nrow(unique(tblLAB_RNA)) > 0) {
  tblLAB_RNA = unique(tblLAB_RNA)
}
```

## `close_d` 

Program level, maximum `close_d` was brought from revised close dates.  

```{r echo=FALSE}
tblLAB_RNA = merge(tblLAB_RNA, program_close_d, by = "program",
                   all.x = TRUE, all.y = FALSE)
```

## Dates

- excluding records with missing dates

```{r echo=FALSE}
frq(tblLAB_RNA, is.na(rna_d))
```

```{r include=FALSE}
tblLAB_RNA = tblLAB_RNA[!is.na(rna_d), ]
```

<!-- - keeping only precise dates (note: `rna_d_a` only available in new data formats) -->

```{r include=FALSE}
# date approximation
# might need processing if values other than `D`!
# frq(tblLAB_RNA, rna_d_a)
tblLAB_RNA[, rna_d_a := NULL]
```

## Exclusion: implausible dates 

- removing tests done after official close dates of the cohort (`max_close_d`)

```{r echo=FALSE}
# View(tblLAB_RNA[rna_d > max_close_d, ])
frq(tblLAB_RNA, rna_d > max_close_d)
```

```{r include=FALSE}
tblLAB_RNA = tblLAB_RNA[!(rna_d > max_close_d), ]
```

- remove tests done before 2003

```{r echo=FALSE}
frq(tblLAB_RNA, rna_d < as.Date("2003-01-01"))
```

```{r include=FALSE}
tblLAB_RNA = tblLAB_RNA[!(rna_d < as.Date("2003-01-01")), ]
```

## Missing lab values

```{r echo=FALSE}
# View(tblLAB_RNA[is.na(rna_v), ])
frq(tblLAB_RNA, is.na(rna_v))
# frq(tblLAB_RNA[is.na(rna_v), ], program)
```

```{r include=FALSE}
tblLAB_RNA = tblLAB_RNA[!is.na(rna_v), ]
```

## Undetectable VL

Replacing undetectable viral load with `detection limit value - 1`, for negative values of `rna_v` use *inverse of that value minus one* instead   

```{r eval=FALSE, include=FALSE}
# see also info on -88 codes
# https://github.com/IeDEA-SA/IeDEA_data-quality/issues/35
frq(tblLAB_RNA[rna_v < 0, ], rna_v)
frq(tblLAB_RNA[rna_v < 0, ], rna_l)
# temp = as.data.frame(table(tblLAB_RNA[rna_v < 0, ]$rna_v, tblLAB_RNA[rna_v < 0, ]$rna_l))
```

```{r}
tblLAB_RNA[, rna_v_orig := rna_v]
tblLAB_RNA[, rna_corr := 0]

# frq(tblLAB_RNA, rna_v == -1 & rna_l > 0 & !is.na(rna_l))
# frq(tblLAB_RNA[rna_v == -1 & rna_l > 0 & !is.na(rna_l), ], rna_l)
tblLAB_RNA[rna_v == -1 & rna_l > 0 & !is.na(rna_l), rna_corr := 1]
tblLAB_RNA[rna_v == -1 & rna_l > 1 & !is.na(rna_l), rna_v := rna_l - 1]
tblLAB_RNA[rna_v == -1 & rna_l == 1 & !is.na(rna_l), rna_v := 1]

# frq(tblLAB_RNA, rna_v == -1 & rna_l == 0)
tblLAB_RNA[rna_v == -1 & rna_l == 0, rna_corr := 1]
tblLAB_RNA[rna_v == -1 & rna_l == 0, rna_v := 1]

# frq(tblLAB_RNA, rna_v == -1 & is.na(rna_l))
tblLAB_RNA[rna_v == -1 & is.na(rna_l), rna_corr := 1]
tblLAB_RNA[rna_v == -1 & is.na(rna_l), rna_v := 1]

# frq(tblLAB_RNA, rna_v < -1 & is.na(rna_l))
# frq(tblLAB_RNA[rna_v < -1 & is.na(rna_l), ], rna_v)
tblLAB_RNA[rna_v < -1 & is.na(rna_l), rna_corr := 1]
tblLAB_RNA[rna_v < -1 & is.na(rna_l), rna_v := (rna_v * -1) - 1]

# frq(tblLAB_RNA, rna_v < -1 & !is.na(rna_l))
# frq(tblLAB_RNA[rna_v < -1 & !is.na(rna_l), ], rna_v)
# frq(tblLAB_RNA[rna_v < -1 & !is.na(rna_l), ], rna_l)
tblLAB_RNA[rna_v < -1 & !is.na(rna_l), rna_corr := 1]
tblLAB_RNA[rna_v < -1 & !is.na(rna_l), rna_v := rna_l - 1]
```

```{r eval=FALSE, include=FALSE}
frq(tblLAB_RNA, rna_corr)
summary(tblLAB_RNA$rna_v)
tblLAB_RNA[, rna_v_orig := NULL]
```

## Multiple daily records

There are patients having 2-3 measurements per day; *geometric* mean of measurements was taken in such cases

```{r echo=FALSE}
tblLAB_RNA[, rna_daily := .N, by = c("patient", "rna_d")]
frq(tblLAB_RNA, rna_daily)
# View(tblLAB_RNA[rna_daily > 1, ])
```

Example:  

```{r echo=FALSE}
tblLAB_RNA[patient == "I_SA_CIDRZ_308621-01" & rna_d == as.Date("2018-09-12"), ][, !c("program", "rna_l", "max_close_d")]
```

```{r include=FALSE}
tblLAB_RNA[rna_daily > 1 & rna_corr == 0, rna_corr := 1]
tblLAB_RNA[rna_daily > 1 & rna_v == 0, rna_v := 1]

tblLAB_RNA[, rna_v := geometric.mean(rna_v), by = c("patient", "rna_d")]
tblLAB_RNA[, rna_order := 1:.N, by = c("patient", "rna_d")]
tblLAB_RNA = tblLAB_RNA[rna_order == 1, ]
tblLAB_RNA[, c("rna_order", "rna_daily") := NULL]
```

## Share of corrected records 

```{r echo=FALSE}
frq(tblLAB_RNA, rna_corr)
```

## Categorical VL

Created two types of categories (rna_v_cat1 & rna_v_cat2).  

```{r eval=FALSE, include=FALSE}
summary(tblLAB_RNA$rna_v)
max(tblLAB_RNA$rna_v)
frq(tblLAB_RNA, is.na(rna_v))
```

```{r include=FALSE}
tblLAB_RNA$rna_v_cat1 = cut(tblLAB_RNA$rna_v,
                            breaks = c(
                              0, 50, 200, 400, 1000,
                              max(tblLAB_RNA$rna_v, na.rm = TRUE) + 1
                            ),
                            labels = c("< 50", "50-199", "200-399", "400-999", ">= 1000"),
                            right = FALSE
)
```

```{r echo=FALSE}
frq(tblLAB_RNA, rna_v_cat1)
```

```{r include=FALSE}
tblLAB_RNA$rna_v_cat2 = cut(tblLAB_RNA$rna_v,
                            breaks = c(
                              0, 50, 1000,
                              max(tblLAB_RNA$rna_v, na.rm = TRUE) + 1
                            ),
                            labels = c("< 50", "50-999", ">= 1000"),
                            right = FALSE
)
```

```{r echo=FALSE}
frq(tblLAB_RNA, rna_v_cat2)
```

## Binary low level viremia indicator

```{r include=FALSE}
tblLAB_RNA$rna_viremia = ifelse(tblLAB_RNA$rna_v >= 50 & tblLAB_RNA$rna_v <= 500,
                                1, 0
)
```

```{r echo=FALSE}
frq(tblLAB_RNA, rna_viremia)
```

## Count and order of lab results

Lab count and order were then calculated  

```{r include=FALSE}
tblLAB_RNA = tblLAB_RNA[, .(
  patient, program,
  rna_d, rna_v,
  rna_v_cat1, rna_v_cat2, rna_viremia
)]

setorder(tblLAB_RNA, patient, rna_d)
tblLAB_RNA[, rna_count := .N, by = patient]
tblLAB_RNA[, rna_order := 1:.N, by = patient]

# frq(tblLAB_RNA$rna_count)
# frq(tblLAB_RNA$rna_order)
```

## Patients without any lab

```{r include=FALSE}
# patients with any art
include = unique(tblLAB_RNA[, .(patient)])

# anti join - so patients without any art here
no_rna = setDT(tblBAS)[!include, on = "patient"]
```

There are `r number(nrow(no_rna), big.mark = ",")` patients from all combined cohorts *without any RNA record*. 

```{r include=FALSE}
rm(include, no_rna); gc()
```

<!-- ----------------------------------------------------- -->

# `tblLAB_CD4`

```{r}
tblLAB_CD4 = data.table(read_fst("../IeDEA_core/data/Stata_202308_Aug/tblLAB_CD4.fst"))
```

Starting point is `r number(nrow(tblLAB_CD4), big.mark = ",")` ART records of `r number(length(unique(tblLAB_CD4$patient)), big.mark = ",")` patients from all combined cohorts.  

## BAS overlap

```{r}
tblLAB_CD4 = merge(tblBAS[, .(patient, program)],
                   tblLAB_CD4,
                   by = c("patient"),
                   all.x = FALSE, all.y = FALSE
)
```

## Keep unique rows

```{r}
# remove duplicate rows
tblLAB_CD4 = distinct(tblLAB_CD4)
```

## `close_d` 

Program level, maximum `close_d` was brought from revised close dates.  

```{r echo=FALSE}
tblLAB_CD4 = merge(tblLAB_CD4, program_close_d, by = "program",
                   all.x = TRUE, all.y = FALSE)
```

## Dates 

- recoding lab values with `cd4_d` bigger than today to `NA`s

```{r echo=FALSE}
# View(tblLAB_CD4[cd4_d >= Sys.Date(), ])
frq(tblLAB_CD4, cd4_d >= Sys.Date())
```

```{r include=FALSE}
tblLAB_CD4[cd4_d >= Sys.Date(), cd4_d := NA]
```

- excluding records with missing dates

```{r echo=FALSE}
frq(tblLAB_CD4, is.na(cd4_d))
```

```{r include=FALSE}
tblLAB_CD4 = tblLAB_CD4[!is.na(cd4_d), ]
```

<!-- - keeping only precise dates (note: `cd4_d_a` only available in new data formats) -->

```{r include=FALSE}
# might need processing if values other than `D`!
# frq(tblLAB_CD4, cd4_d_a)
tblLAB_CD4[, cd4_d_a := NULL]
```

## Exclusion: implausible dates 

- removing tests done after official close dates of the cohort (`max_close_d`)

```{r echo=FALSE}
# View(tblLAB_CD4[cd4_d > max_close_d, ])
frq(tblLAB_CD4, cd4_d > max_close_d)
```

```{r include=FALSE}
tblLAB_CD4 = tblLAB_CD4[!(cd4_d > max_close_d), ]
```

- remove tests done before 2003

```{r echo=FALSE}
frq(tblLAB_CD4, cd4_d < as.Date("2003-01-01"))
```

```{r include=FALSE}
tblLAB_CD4 = tblLAB_CD4[cd4_d >= as.Date("2003-01-01"), ]
```

## Missing or implausible values

- missing lab values

```{r echo=FALSE}
# View(tblLAB_CD4[is.na(cd4_v), ])
frq(tblLAB_CD4, is.na(cd4_v))
```

```{r include=FALSE}
tblLAB_CD4 = tblLAB_CD4[!is.na(cd4_v), ]
```

- below `0` 

```{r echo=FALSE}
# View(tblLAB_CD4[cd4_v < 0, ])
frq(tblLAB_CD4, cd4_v < 0)
```

```{r include=FALSE}
tblLAB_CD4 = tblLAB_CD4[!(cd4_v < 0), ]
```

## Multiple daily records

There are patients having 2-3 measurements per day; mean of measurements was taken in such cases  

```{r echo=FALSE}
tblLAB_CD4[, cd4_daily := .N, by = c("patient", "cd4_d")]
frq(tblLAB_CD4, cd4_daily)
```

```{r include=FALSE}
# View(tblLAB_CD4[cd4_daily > 1, ])
# tblLAB_CD4[patient == "KI06000017459", ] # (sic!)

tblLAB_CD4[, cd4_v := mean(cd4_v), by = c("patient", "cd4_d")]
tblLAB_CD4[, cd4_order := 1:.N, by = c("patient", "cd4_d")]
tblLAB_CD4 = tblLAB_CD4[cd4_order == 1, ]
tblLAB_CD4[, c("cd4_daily", "cd4_order") := NULL]
```

## Count and order of CD4 results

Lab count and order were then calculated  

```{r include=FALSE}
tblLAB_CD4 = tblLAB_CD4[, .(patient, program, cd4_d, cd4_v, cd4_u)]

# keep unique rows
# nrow(tblLAB_CD4) - nrow(unique(tblLAB_CD4))
# tblLAB_CD4 = unique(tblLAB_CD4)

setorder(tblLAB_CD4, patient, cd4_d)
tblLAB_CD4[, cd4_count := .N, by = patient]
tblLAB_CD4[, cd4_order := 1:.N, by = patient]

# frq(tblLAB_CD4$rna_count)
# frq(tblLAB_CD4$cd4_order)
```

## Patients without any lab

```{r include=FALSE}
# patients with any art
include = unique(tblLAB_CD4[, .(patient)])

# anti join - so patients without any art here
no_cd4 = setDT(tblBAS)[!include, on = "patient"]
```

There are `r number(nrow(no_cd4), big.mark = ",")` patients from all combined cohorts *without any CD4 record*. 

```{r include=FALSE}
rm(include, no_cd4); gc()
```



<!-- ----------------------------------------------------- -->

# Saving data 

## Local  

This state of data is preserved in `data` directory. `fst` files for R and `dta` files for Stata are saved.    

```{r}
write_fst(tblBAS, "data_temp/tblBAS_v003.fst")
# tblBAS = data.table(read_fst("data_temp/tblBAS_v003.fst"))

write_fst(tblLTFU, "data_temp/tblLTFU_v003.fst")
# tblLTFU = data.table(read_fst("data_temp/tblLTFU_v003.fst"))

write_fst(tblART, "data_temp/tblART_v003.fst")
# tblART = data.table(read_fst("data_temp/tblART_v003.fst"))

write_fst(tblLAB_RNA, "data_temp/tblLAB_RNA_v003.fst")
# tblLAB_RNA = data.table(read_fst("data_temp/tblLAB_RNA_v003.fst"))

write_fst(tblLAB_CD4, "data_temp/tblLAB_CD4_v003.fst")
# tblLAB_CD4 = data.table(read_fst("data_temp/tblLAB_CD4_v003.fst"))

```

<!-- ----------------------------------------------------- -->

# Content & missings

## `tblBAS`

```{r echo=FALSE}
skimr::skim(tblBAS[,.SD, .SDcols = !c("age_cat1", "age_cat2")])
```

## `tblLAB_CD4`

```{r echo=FALSE}
skimr::skim(tblLAB_CD4)
```

## `tblLAB_RNA`

```{r echo=FALSE}
skimr::skim(tblLAB_RNA)
```

<!-- ----------------------------------------------------- -->

# Computing Environment

```{r echo=FALSE, results='asis'}
report::report(sessionInfo())
```

<!-- ----------------------------------------------------- -->

