---
title: "SA256 DTG uptake <br> tblVIS"
subtitle: " "
author: "Radek Panczak, Eliane Rohner, Elizabeth Zaniewski"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    code_folding: show
    self_contained: true
    highlight: pygments
    keep_md: no
    toc_depth: 3
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
---

```{r r-setup, include = FALSE}
options(scipen = 999)
options(max.print = "75")
set.seed(12345)

library(pacman)
p_load(
  kableExtra,
  scales, ggplot2, dplyr,
  fst, data.table, sjmisc,
  osfr
)

import::from("psych", "geometric.mean")
import::from("sjmisc", "frq")
import::from("gmodels", "CrossTable")
```

```{r conflicts, include = FALSE}
conflicted::conflict_scout()
```

```{r knit-setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

knitr::opts_chunk$set(
  cache = FALSE,
  prompt = FALSE,
  tidy = FALSE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  echo = TRUE
)

knitr::opts_knit$set(width = 75)
```

<!-- ----------------------------------------------------- -->


```{r}
#Clear existing data and graphics
rm(list=ls())
graphics.off()
```



# IeDEA-SA Data Prep

Raw Stata and `fst` files come from central directory of `IeDEA_core`.  

Using release `Stata_202308_Aug`.  

```{r eval=FALSE, include=FALSE}
list.files("../IeDEA_core/data/Stata_202308_Aug/", pattern = "\\.fst$") %>%
  kable(col.names = "Datasets - new") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = F
  )
```

## Druglist

Updated `druglist` dataset from the `ART-codes` repo.   

```{r include=FALSE}
druglist = data.table(readr::read_csv("https://raw.githubusercontent.com/IeDEA-SA/ART-codes/main/data-raw/IeDEA_druglist.csv"))[, c("drug_orig") := NULL]

druglist[, drug := trimws(drug, which = "both")]
```

```{r echo=FALSE}
druglist %>%
  kbl() %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  column_spec(1, italic = TRUE) %>%
  row_spec(1, italic = FALSE)
```

```{r include=FALSE}
druglist = druglist[, .(art_id, drug, arv_class)]
```

## Official DTG adoption dates

Pulled by @aezaniewski from different sources.  

```{r include=FALSE}
dtg_recomend = data.table(read_fst("data_temp/dtg_recomend_v001.fst"))
```

```{r echo=FALSE}
dtg_recomend %>%
  kbl() %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  column_spec(1, italic = TRUE) %>%
  row_spec(1, italic = FALSE)
```

```{r include=FALSE}

rev_close_d = data.table(read_fst("data_temp/rev_close_d_v001.fst"))

program_close_d = data.table(read_fst("data_temp/program_close_d_v001.fst"))

tblCENTER = data.table(read_fst("data_temp/tblCENTER_v001.fst"))

tblVIS = data.table(read_fst("data_temp/tblVIS_v001.fst"))

tblBAS = data.table(read_fst("data_temp/tblBAS_v003.fst"))

tblLTFU = data.table(read_fst("data_temp/tblLTFU_v003.fst"))

tblART = data.table(read_fst("data_temp/tblART_v003.fst"))

tblLAB_RNA = data.table(read_fst("data_temp/tblLAB_RNA_v003.fst"))

tblLAB_CD4 = data.table(read_fst("data_temp/tblLAB_CD4_v003.fst"))

```



<!-- ----------------------------------------------------- -->

# `tblVIS`

Starting point is `r number(nrow(tblVIS), big.mark = ",")` ART records of `r number(length(unique(tblVIS$patient)), big.mark = ",")` patients from all combined cohorts.  


Keep unique rows

There are few surplus duplicates; keeping only unique 

```{r}
tblVIS = distinct(tblVIS)
```


Find program max `close_d` 

Program level, maximum `close_d` was brought from revised close dates.  

```{r echo=FALSE}
tblVIS = merge(tblVIS, program_close_d, by = "program",
                   all.x = TRUE, all.y = FALSE)
```


Drop implausible or missing visit dates 

```{r eval=FALSE, include=FALSE}
frq(tblVIS, is.na(vis_d))
tblVIS = tblVIS[!is.na(vis_d), ]
```

Keeping only precise dates (don't do)

```{r echo=FALSE}
frq(tblVIS, vis_d_a)
```

```{r include=FALSE}
# View(tblVIS[vis_d_a == "", ])
#tblVIS = tblVIS[vis_d_a == "D", ]
#tblVIS[, vis_d_a := NULL]
```

Remove visits before 2003

```{r echo=FALSE}
frq(tblVIS, vis_d < as.Date("2003-01-01"))
```

```{r include=FALSE}
tblVIS = tblVIS[vis_d >= as.Date("2003-01-01"), ]
```


## Check that visit center names match tblCENTER

Compare tblCENTER and tblVIS center names of patient's last visit

```{r}
name_tblCENTER = tblCENTER[, .(center, program, country)]
name_tblCENTER = name_tblCENTER[, center_center := center]
setnames(name_tblCENTER, "program", "center_program")
setnames(name_tblCENTER, "country", "center_country")

name_tblCENTER = distinct(name_tblCENTER)
frq(name_tblCENTER, center)
```


```{r}
name_tblVIS = tblVIS[, .(patient, vis_d, center, program, country)]
name_tblVIS = distinct(name_tblVIS)

setDT(name_tblVIS)

setorder(name_tblVIS, patient, vis_d, center, program, country)

name_tblVIS = name_tblVIS[, vis_count := .N, by = patient]


name_tblVIS = name_tblVIS[, vis_order := 1:.N, by = patient]

name_tblVIS = name_tblVIS[vis_count == vis_order,]

name_tblVIS = name_tblVIS[, last_vis_d := vis_d]


name_tblVIS = name_tblVIS[, vis_center := center]
setnames(name_tblVIS, "program", "vis_program")
setnames(name_tblVIS, "country", "vis_country")
name_tblVIS = distinct(name_tblVIS)
frq(name_tblVIS, center)
```


Program centers in tblCENTER not in tblVIS

```{r echo=FALSE}
name_match_CENTER = merge(name_tblCENTER, name_tblVIS, by = "center",
                   all.x = TRUE, all.y = TRUE)
name_match_CENTER = name_match_CENTER[is.na(vis_program), ]

frq(name_match_CENTER, center_program)
frq(name_match_CENTER, center_center)
```


Program centers in tblVIS not in tblCENTER

```{r echo=FALSE}
name_match_VIS = merge(name_tblVIS, name_tblCENTER, by = "center",
                   all.x = TRUE, all.y = TRUE)
name_match_VIS = name_match_VIS[is.na(center_program), ]

frq(name_match_VIS, vis_program)
frq(name_match_VIS, vis_center)

```

Correct any center names in tblVIS if they are spelled differently than in tblCENTER (none)

```{r}
#tblVIS[, center := ifelse(center =="AFRICA CENTER", "AFRICA CENTRE", center)]
#tblVIS[, center := ifelse(center =="TLC-SA", "THEMBA LETHU CLINIC", center)]
```

## Exclude patients with last visit at centers not in tblCENTER

```{r echo=FALSE}
exclude = name_match_VIS[, drop :=1,]

exclude = exclude[drop ==1,. (patient, vis_program, vis_center, drop)]

exclude = unique(exclude)

frq(exclude, vis_program)

```


```{r}
exclude = exclude[,. (patient)]
```


Exclude patients across tables 

```{r}

tblBAS = setDT(tblBAS)[!exclude, on = c("patient")]

tblVIS = setDT(tblVIS)[!exclude, on = c("patient")]
tblART = setDT(tblART)[!exclude, on = c("patient")]
tblLAB_CD4 = setDT(tblLAB_CD4)[!exclude, on = c("patient")]
tblLAB_RNA = setDT(tblLAB_RNA)[!exclude, on = c("patient")]
tblLTFU = setDT(tblLTFU)[!exclude, on = c("patient")]

```


```{r include=FALSE}
rm(name_match_CENTER, name_match_VIS, name_tblCENTER, name_tblVIS); gc()
```


## BAS overlap

Keeping only tblVIS patients that are also in tblBAS

```{r}
temp = tblBAS[, .(patient, recart_d, dtg_recomend_d)]

temp = distinct(temp)

tblVIS = merge(tblVIS, temp,
               by = c("patient"),
               all.x = FALSE, all.y = FALSE
)

```

## tblVIS & tblBAS: update max_close_d for patients with last visit at dropped center

Find center where patient had last visit

```{r include=FALSE}
center = tblCENTER[, .(center, add_center_d, drop_center_d)]
```


```{r include=FALSE}
tblVIS = merge(tblVIS, center,
               by = c("center"),
               all.x = FALSE, all.y = FALSE
)

```


```{r include=FALSE}
setorder(tblVIS, patient, vis_d)
tblVIS[, vis_count := .N, by = patient]
tblVIS[, vis_order := 1:.N, by = patient]

last_visit = tblVIS[vis_count == vis_order,]

last_visit = last_visit[, last_vis_d := vis_d]

last_visit = last_visit[, .(patient, program, country, max_close_d, recart_d, last_vis_d, center, add_center_d, drop_center_d, dtg_recomend_d, vis_count)]

```

```{r eval=FALSE, include=FALSE}
test <- last_visit %>% filter(country =="ZAF")
```


Find patients who had last visit at center dropped before max_close_d of program

```{r include=FALSE}

last_visit = last_visit[!is.na(drop_center_d) & drop_center_d < max_close_d, drop := 1]

last_visit = last_visit[drop == 1,]

```

tblBAS: update max_close_d to the drop_center_d at last visit

```{r}
tblBAS = merge(tblBAS, last_visit[, .(patient, drop_center_d)],
               by = c("patient"),
               all.x = TRUE, all.y = FALSE
)

setDT(tblBAS)

tblBAS = tblBAS[!is.na(drop_center_d), max_close_d := drop_center_d]

test <- tblBAS %>% filter(max_close_d == drop_center_d)


```


tblVIS: update max_close_d to the drop_center_d at last visit

```{r}

tblVIS = tblVIS[!is.na(drop_center_d), max_close_d := drop_center_d]

```


## Exclusion: close date (program)

Programs with visits after close date (don't need to do)

```{r}
#frq(tblVIS, vis_d > max_close_d)
frq(tblVIS[vis_d > max_close_d, ], program)
```


Remove visits after close date of the cohort/center (`max_close_d`) (don't need to do)

```{r}
#tblVIS = tblVIS[vis_d <= max_close_d, ]
#tblVIS[, max_close_d := NULL]
```



## Implausible next visit dates 

- next visit estimates that are the same or earlier than the visit are replaced to NAs  

```{r echo=FALSE}
frq(tblVIS, !is.na(next_visit_d) & next_visit_d <= vis_d)
test <- tblVIS %>% filter(!is.na(next_visit_d) & next_visit_d <= vis_d)
```

```{r include=FALSE}
tblVIS = tblVIS[!is.na(next_visit_d) & next_visit_d <= vis_d, next_visit_d_a := "U"]
tblVIS = tblVIS[!is.na(next_visit_d) & next_visit_d <= vis_d, next_visit_d := NA]
```

- cleaning up imprecise next visit estimates  

```{r echo=FALSE}
frq(tblVIS, next_visit_d_a)
```

```{r include=FALSE}
#tblVIS[next_visit_d_a %in% c("Y", "U", ""), next_visit_d := NA]
tblVIS = tblVIS[, next_visit_d_a := NA]
```

## Multiple visits per day  

There are few patients having 2-3 visits per day

```{r echo=FALSE}
tblVIS[, vis_daily := .N, by = c("patient", "vis_d")]
frq(tblVIS, vis_daily)
# View(tblVIS[vis_daily > 1, ])
```

```{r include=FALSE}
tblVIS[, vis_daily := NULL]
```

## Count and order of visits

Visit count and order were calculated on raw data

```{r include=FALSE}
setorder(tblVIS, patient, vis_d)
tblVIS[, vis_count := .N, by = patient]
tblVIS[, vis_order := 1:.N, by = patient]

# frq(tblVIS$vis_count)
# frq(tblVIS$vis_order)
```

## Patients without any visits

```{r include=FALSE}
# patients with any art
include = unique(tblVIS[, .(patient)])

# anti join - so patients without any art here
no_vis = setDT(tblBAS)[!include, on = "patient"]

```

There are `r number(nrow(no_vis), big.mark = ",")` patients from all combined cohorts *without any visit*. 

```{r echo=FALSE}
frq(no_vis, program)
```



```{r eval=FALSE, include=FALSE}
# all those patients have at least one record somewhere else though
temp1 = unique(rbind(unique(tblART[, .(patient)]), 
                     unique(tblLAB_CD4[, .(patient)]), 
                     unique(tblLAB_RNA[, .(patient)])))

temp2 = setDT(tblBAS)[!temp1, on = c("patient")]

rm(temp1, temp2); gc()
```

## Drop patients in tblBAS without visits in tblVIS

```{r}
exclude = no_vis[,. (patient)]
```


Exclude patients across tables 

```{r}

tblBAS = setDT(tblBAS)[!exclude, on = c("patient")]
tblVIS = setDT(tblVIS)[!exclude, on = c("patient")]
tblART = setDT(tblART)[!exclude, on = c("patient")]
tblLAB_CD4 = setDT(tblLAB_CD4)[!exclude, on = c("patient")]
tblLAB_RNA = setDT(tblLAB_RNA)[!exclude, on = c("patient")]
tblLTFU = setDT(tblLTFU)[!exclude, on = c("patient")]

```


```{r include=FALSE}
rm(include, no_vis, exclude, temp, test, center); gc()
```


<!-- ----------------------------------------------------- -->

# Last Clinical Encounter

## Calculate 

Last visit date (*if available!*)  

```{r}
setorder(tblVIS, "patient", "vis_d")

tblBAS = tblVIS[, .(patient, last_vis_d = vis_d)][tblBAS, on = "patient", mult = "last"]
```


Collecting all encounters from four datasets    

```{r}
tblCLIN = rbind(
  tblART[, .(patient, enc_d = art_sd)],
  tblLAB_CD4[, .(patient, enc_d = cd4_d)],
  tblLAB_RNA[, .(patient, enc_d = rna_d)],
  tblVIS[, .(patient, enc_d = vis_d)]
)


tblCLIN = unique(tblCLIN, by = c("patient", "enc_d"))

tblCLIN = merge(tblCLIN, tblBAS[, .(patient, max_close_d)],
               by = c("patient"),
               all.x = TRUE, all.y = FALSE
)

tblCLIN = tblCLIN[enc_d <= max_close_d,]


stopifnot(tblCLIN[, all(!is.na(enc_d))])
stopifnot(tblCLIN[, all(!is.na(enc_d))])

setorder(tblCLIN, "patient", "enc_d")
```


Getting *last clinical encounter* for each patient on the basis of all dated events from ART, CD4, RNA & VIS tables   

```{r}
tblBAS = tblCLIN[, .(patient, last_enc_d = enc_d)][tblBAS, on = "patient", mult = "last"]
```

Patient with a record in tblCLIN

```{r}
frq(tblBAS, is.na(last_enc_d))
```


Drop patients without a record in tblCLIN

```{r}
exclude = tblBAS[is.na(last_enc_d),. (patient)]
```


Exclude patients across tables 

```{r}

tblBAS = setDT(tblBAS)[!exclude, on = c("patient")]
tblVIS = setDT(tblVIS)[!exclude, on = c("patient")]
tblART = setDT(tblART)[!exclude, on = c("patient")]
tblLAB_CD4 = setDT(tblLAB_CD4)[!exclude, on = c("patient")]
tblLAB_RNA = setDT(tblLAB_RNA)[!exclude, on = c("patient")]
tblLTFU = setDT(tblLTFU)[!exclude, on = c("patient")]

```


Last encounter has advantage of being complete...  

```{r echo=FALSE}
frq(tblBAS, is.na(last_enc_d))
```

... when compared to last visit only:  

```{r echo=FALSE}
frq(tblBAS, is.na(last_vis_d))
```


Last encounter by program:  

```{r echo=FALSE}
ggplot(
  tblBAS,
  aes(x = last_enc_d)
) +
  geom_density() +
  facet_wrap(vars(program)) +
  scale_x_date() +
  theme_minimal() +
  xlab("Last encounter day")
```


## Exclusion: LCE before DTG implementation date

**Patients whose last clinical encounter took place before the DTG implementation date were excluded.**  

```{r echo=FALSE}
# frq(tblBAS, is.na(last_enc_d))
# frq(tblBAS, is.na(dtg_min_program))

frq(tblBAS, last_enc_d < dtg_recomend_d)
```

**Note:** *correctly there are very few DTGs here!*  

```{r echo=FALSE}
frq(tblBAS[last_enc_d < dtg_recomend_d, ], !is.na(dtg_min_pat))
```

```{r include=FALSE}
tblBAS = tblBAS[last_enc_d >= dtg_recomend_d, ]

# exclusion needs to reapplied
tblCLIN = setDT(tblCLIN)[tblBAS[, .(patient)], on = "patient"]

tblLTFU = setDT(tblLTFU)[tblBAS[, .(patient)], on = "patient"]

tblART = setDT(tblART)[tblBAS[, .(patient)], on = "patient"]

tblLAB_CD4 = setDT(tblLAB_CD4)[tblBAS[, .(patient)], on = "patient"]
tblLAB_RNA = setDT(tblLAB_RNA)[tblBAS[, .(patient)], on = "patient"]
tblVIS = setDT(tblVIS)[tblBAS[, .(patient)], on = "patient"]

gc()
```

Last encounter by program after exclusion:  

```{r echo=FALSE}
ggplot(
  tblBAS,
  aes(x = last_enc_d)
) +
  geom_density() +
  facet_wrap(vars(program), scales = "free_y") +
  scale_x_date() +
  theme_minimal() +
  xlab("Last encounter day")
```

Range of values for last visit:  

```{r}
last_visit_range = tblBAS[, .(lower = min(last_vis_d, na.rm = TRUE), upper = max(last_vis_d, na.rm = TRUE)), by = "program"]
```

```{r echo=FALSE}
last_visit_range %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```

And for last encounter:  

```{r}
last_encounter_range = tblBAS[, .(lower = min(last_enc_d, na.rm = TRUE), upper = max(last_enc_d, na.rm = TRUE)), by = "program"]
```

```{r echo=FALSE}
last_encounter_range %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```

```{r include=FALSE}
rm(last_visit_range, last_encounter_range); gc()
```




<!-- ----------------------------------------------------- -->

# Saving data 

## Local  

This state of data is preserved in `data` directory. `fst` files for R and `dta` files for Stata are saved.    

```{r}
write_fst(tblBAS, "data_temp/tblBAS_v004.fst")
# tblBAS = data.table(read_fst("data_temp/tblBAS_v004.fst"))

write_fst(tblLTFU, "data_temp/tblLTFU_v004.fst")
# tblLTFU = data.table(read_fst("data_temp/tblLTFU_v004.fst"))

write_fst(tblART, "data_temp/tblART_v004.fst")
# tblART = data.table(read_fst("data_temp/tblART_v004.fst"))

write_fst(tblLAB_RNA, "data_temp/tblLAB_RNA_v004.fst")
# tblLAB_RNA = data.table(read_fst("data_temp/tblLAB_RNA_v004.fst"))

write_fst(tblLAB_CD4, "data_temp/tblLAB_CD4_v004.fst")
# tblLAB_CD4 = data.table(read_fst("data_temp/tblLAB_CD4_v004.fst"))

write_fst(tblVIS, "data_temp/tblVIS_v004.fst")
# tblVIS = data.table(read_fst("data_temp/tblVIS_v004.fst"))

write_fst(tblCLIN, "data_temp/tblCLIN_v004.fst")
# tblCLIN = data.table(read_fst("data_temp/tblCLIN_v004.fst"))
```

<!-- ----------------------------------------------------- -->

# Content & missings

## `tblBAS`

```{r echo=FALSE}
skimr::skim(tblBAS[,.SD, .SDcols = !c("age_cat1", "age_cat2")])
```

```{r echo=FALSE, message=FALSE}
visdat::vis_miss(tblBAS[,.SD, .SDcols = !c("patient", "program", "age_cat1", "age_cat2")],
                 warn_large_data = FALSE)
```

## `tblVIS`

```{r echo=FALSE}
skimr::skim(tblVIS)
```

<!-- ----------------------------------------------------- -->

# Computing Environment

```{r echo=FALSE, results='asis'}
report::report(sessionInfo())
```

<!-- ----------------------------------------------------- -->

