---
title: "SA256 DTG uptake <br> IWHOD2024_Poster_Fig1_cohorts"
subtitle: " "
author: "Elizabeth Zaniewski,Radek Panczak, Eliane Rohner"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    code_folding: show
    self_contained: true
    highlight: pygments
    keep_md: no
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
---

```{r r-setup, include = FALSE}
options(scipen = 999)
options(max.print = "300")
set.seed(12345)

library(pacman)
p_load(sf, tidyverse, kableExtra, report, magrittr, fst, data.table, survival, lubridate, ggsurvfit, gtsummary, tidycmprsk, Hmisc, survminer, survMisc)

p_load(
  kableExtra,
  scales, ggplot2, dplyr,
  fst, data.table, sjmisc,
  osfr
)

#install.packages("finalfit")

library(survival)
library(lubridate)
library(ggsurvfit)
library(gtsummary)
library(tidycmprsk)
library(Hmisc)
library(survminer)
library(forcats)
library(dplyr)
#library(condsurv)
library(patchwork)
library(ggplot2)
library(cowplot)



import::from("sjmisc", "frq")
import::from("gmodels", "CrossTable")
```

```{r conflicts, include = FALSE}
tidyverse::tidyverse_conflicts()

conflicted::conflict_scout()
```

```{r knit-setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

knitr::opts_chunk$set(cache = FALSE,
                      prompt = FALSE,
                      tidy = FALSE,
                      comment = NA,
                      message = FALSE,
                      warning = FALSE,
                      echo = FALSE)

knitr::opts_knit$set(width = 200)
```

<!-- ----------------------------------------------------- -->

```{r}
#Clear existing data and graphics
rm(list=ls())
graphics.off()
```

```{r}

#####################################################################
# to adjust decimal places of categorical variables

my.render.cat <- function(x) {
  c("", sapply(stats.default(x), function(y) with(y,
                                                  sprintf("%d (%0.0f %%)", FREQ, PCT))))
}

#####################################################################

```


<!-- ----------------------------------------------------- -->


# National DTG adoption dates

Pulled by @aezaniewski from different sources.  

```{r include=FALSE}
dtg_recomend = data.table(read_fst("data_temp/dtg_recomend_v001.fst"))

dtg_recomend$ctry <- factor(dtg_recomend$country, levels = c( "LSO","MWI", "MOZ" ,"ZAF", "ZMB", "ZWE"),
 labels = c("Lesotho",  "Malawi", "Mozambique", "South Africa", "Zambia", "Zimbabwe"))



```

```{r echo=FALSE}
dtg_recomend %>% select(country, ctry, dtg_recomend_d_initiate, dtg_recomend_d) %>%
  kbl() %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  column_spec(1, italic = TRUE) %>%
  row_spec(1, italic = FALSE)
```

<!-- ----------------------------------------------------- -->

# IWHOD 2024 Abstract


## Title

**Trends in dolutegravir uptake by sex in Southern Africa**


<!-- ----------------------------------------------------- -->


# Load wrangled data: Initiating


```{r include=FALSE}
tblBAS_1 = data.table(read_fst("data_temp/tblBAS_cohort1_LTFU_WHO_v010.fst"))

tblBAS_1$ctry <- factor(tblBAS_1$country, levels = c( "LSO","MWI", "MOZ" ,"ZAF", "ZMB", "ZWE"),
 labels = c("Lesotho",  "Malawi", "Mozambique", "South Africa", "Zambia", "Zimbabwe"))


```


```{r include=FALSE}

# Create sex and age categories
tblBAS_1 <- tblBAS_1 %>%
  mutate(sex_age2_txt = ifelse(sex_age2 =="Male.[20,50)", "Male 20-49",
                               ifelse(sex_age2 =="Female.[20,50)", "Female 20-49",
                                    ifelse(sex_age2 =="Male.[50,100]", "Male 50+",
                                         ifelse(sex_age2 =="Female.[50,100]", "Female 50+", "other"
                                      )))))

tblBAS_1 <- tblBAS_1 %>%
  mutate(sex_age3_txt = ifelse(sex_age3 =="Male.[20,30)", "Male 20-29",
                               ifelse(sex_age3 =="Female.[20,30)", "Female 20-29",
                                ifelse(sex_age3 =="Male.[30,40)", "Male 30-39",
                               ifelse(sex_age3 =="Female.[30,40)", "Female 30-39",
                              ifelse(sex_age3 =="Male.[40,50)", "Male 40-49",
                               ifelse(sex_age3 =="Female.[40,50)", "Female 40-49",
                                ifelse(sex_age3 =="Male.[50,100]", "Male 50+",
                                ifelse(sex_age3 =="Female.[50,100]", "Female 50+", "other"
                                )))))))))

```



```{r echo=FALSE}
table1::label(tblBAS_1$sex) <- "Sex"
table1::label(tblBAS_1$age_cat3) <- "Age group"
table1::label(tblBAS_1$dtg_first_txt) <- "ART at initiation"
table1::label(tblBAS_1$sex_age2) <- "Sex & age group"
table1::label(tblBAS_1$sex_age3) <- "Sex & age group"
table1::label(tblBAS_1$program) <- "Program"
table1::label(tblBAS_1$ctry) <- "Country"
table1::label(tblBAS_1$sex_age2_txt) <- "Sex & age group"
table1::label(tblBAS_1$sex_age3_txt) <- "Sex & age group"

```


## Adults aged 20-39 years only


```{r include=FALSE}
temp_tblBAS_1 <- tblBAS_1 %>% filter(!program =="NEWLANDS") %>% mutate(age2030 = ifelse(age_cat3 == "[20,30)" | age_cat3 =="[30,40)", 1, 0)) %>% mutate(age2030_txt = ifelse(age2030 == 1, "aged 20-39", "aged 40+"))
#frq(temp_tblBAS_1, age2030_txt)
```



```{r include=FALSE}
#All adult data
sjPlot::tab_xtab(temp_tblBAS_1$program,
                 temp_tblBAS_1$age2030_txt,
                 show.row.prc = TRUE, show.summary = FALSE)
```



```{r}
tblBAS_1 <- tblBAS_1 %>% filter(age_cat3 == "[20,30)" | age_cat3 =="[30,40)") %>% filter(!program =="NEWLANDS") 

```



```{r eval=FALSE, include=FALSE}
#Only adults aged 20-39 years
sjPlot::tab_xtab(tblBAS_1$program,
                 tblBAS_1$sex_age3_txt,
                 show.row.prc = TRUE, show.summary = FALSE)
```



```{r}

frq(tblBAS_1, sex)

temp1 <- tblBAS_1 %>% select(country, program, dtg_recomend_d, max_close_d) %>%
  filter(!program =="NEWLANDS") %>%
  arrange(dtg_recomend_d, country, program) %>%
  group_by(country, program, dtg_recomend_d, max_close_d) %>%
  mutate(Npatients = n()) %>%
  distinct()
  
temp1 %>% 
  kbl() %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  column_spec(1, italic = TRUE) %>%
  row_spec(1, italic = FALSE)

```


## Overview by country

```{r}
sjPlot::tab_xtab(tblBAS_1$ctry,
                 tblBAS_1$sex,
                 show.row.prc = TRUE, show.summary = FALSE)
```


```{r}
sjPlot::tab_xtab(tblBAS_1$ctry,
                 tblBAS_1$sex_age3_txt,
                 show.row.prc = TRUE, show.summary = FALSE)
```

## Overview by program

```{r}
sjPlot::tab_xtab(tblBAS_1$program,
                 tblBAS_1$sex,
                 show.row.prc = TRUE, show.summary = FALSE)
```


## Initiating "Other ART" vs "DTG ART".

by country

```{r}
sjPlot::tab_xtab(tblBAS_1$ctry,
                 tblBAS_1$dtg_first_txt,
                 show.row.prc = TRUE, show.summary = FALSE)
```

by sex & age

```{r}
sjPlot::tab_xtab(tblBAS_1$sex_age3_txt,
                 tblBAS_1$dtg_first_txt,
                 show.row.prc = TRUE, show.summary = FALSE)
```



## Pregnancy data for South Africa


```{r}
tblVIS = data.table(read_fst("data_temp/tblVIS_v005_LTFU_WHO.fst"))
```

Patient data by cohort and sex

```{r}
sjPlot::tab_xtab(tblBAS_1$program,
                 tblBAS_1$sex,
                 show.row.prc = TRUE, show.summary = FALSE)
```



```{r}
temp_tblBAS_1 = tblBAS_1[sex =="Female", ]

tblBAS_preg = merge(temp_tblBAS_1, 
                  tblVIS[, .(patient, center, vis_d, preg_y)], 
                  by = "patient",
                  all.x = TRUE, all.y = FALSE)

```



```{r include=FALSE}
# Pregnancy data by cohort for women only

label(tblBAS_preg$preg_y) <- "Pregnancy no/yes/unknown"

sjPlot::tab_xtab(tblBAS_preg$program,
                 tblBAS_preg$preg_y,
                 show.row.prc = TRUE, show.summary = FALSE)
```

Keep South African data for pregnancy within +/- 9 months from ART start

```{r}

tblBAS_preg = tblBAS_preg[country =="ZAF", ]

tblBAS_preg <- tblBAS_preg %>% select(patient, sex, program, country, recart_d, dtg_first_txt, dtg_min_pat, vis_d,  preg_y)

```


```{r}
tblBAS_preg2 <- tblBAS_preg %>% 
  mutate(recart_vis = abs(recart_d - vis_d)) %>% 
  mutate(recart_vis_time = ifelse(recart_vis < 281, 1, 0)) %>% 
  filter(recart_vis_time ==1 & preg_y ==1) %>%
  filter(sex =="Female") %>%
  mutate(preg = 1) %>%
  select(patient, preg, preg_y) %>%
  unique()


```


```{r include=FALSE}


tblBAS_1 = merge(tblBAS_1, 
                  tblBAS_preg2[, .(patient, preg, preg_y)], 
                  by = "patient",
                  all.x = TRUE, all.y = FALSE)

tblBAS_1 = tblBAS_1[is.na(preg), preg2 :=0]

tblBAS_1 = tblBAS_1[!is.na(preg), preg2 :=1]

label(tblBAS_1$preg2) <- "Pregnancy no or unknown/yes"

frq(tblBAS_1, preg2)

```



```{r}
# Pregnancy data 
table1::label(tblBAS_1$preg2) <- "Known pregnancy within +/- 9months of ART start"


```


```{r eval=FALSE, include=FALSE}
sjPlot::tab_xtab(tblBAS_1$program,
                 tblBAS_1$preg2,
                 show.row.prc = TRUE, show.summary = FALSE)
```



HLABISA women only
```{r}
temp_H <- tblBAS_1 %>% filter(program == "HLABISA")


sjPlot::tab_xtab(temp_H$preg2,
                 temp_H$dtg_first_txt,
                 show.row.prc = TRUE, show.summary = FALSE)
```

```{r}
sjPlot::tab_xtab(temp_H$dtg_first_txt,
                 temp_H$preg2,
                 show.row.prc = TRUE, show.summary = FALSE)
```


Histogram
```{r}
temp <- temp_H %>% mutate(status = ifelse(preg2 == 1, "Known pregnancy", "No known pregnancy")) 

ggplot(data = temp, aes(recart_d)) +
  geom_histogram(aes(fill = status), position = "dodge") +
  scale_y_sqrt() 
```


```{r}

ggplot(data = temp, aes(recart_d)) +
  geom_histogram(aes(fill = status), position = "dodge") +
  facet_grid(vars(status))
```




<!-- ----------------------------------------------------- -->

# Figures: Initiating cohort

## All programs

Cumulative incidence (1-KM plots) of intiating DTG by country and sex over calendar year.
(aggregated quarterly)

Adults aged 20 to 39 years at ART initiation.

Using facet_wrap.

```{r}

my_comma <- scales::label_comma(accuracy = 1, big.mark = ",")

PatperProg_1 <- tblBAS_1 %>% 
  group_by(program) %>%
  summarise("NperProg"=n()) %>%
  mutate(NperProg2 = my_comma(NperProg)) %>%
  mutate(text1 = " (n=") %>%
  mutate(text2 = ")") %>%
  mutate(NProg = paste(program, "\n", text1, NperProg2, text2, sep= "")) %>%
  select(program, NProg) 

tblBAS_1 <- merge(tblBAS_1, PatperProg_1, by = "program", 
           all.x = TRUE, all.y = FALSE) 

```




```{r include=FALSE}
tblBAS <- tblBAS_1 %>%
 #filter(recart_d < "2023-08-01") %>%
  mutate(recart_date2 = floor_date(as_date(recart_d), "month")) %>%
  mutate(recart_date = ceiling_date(as_date(recart_d), "month")) %>%
  mutate(recommend_date = floor_date(as_date(dtg_recomend_d), "month")) %>%
  mutate(recommend_qtr = paste0(substring(year(dtg_recomend_d),3,4),"/0",quarter(dtg_recomend_d))) %>%
  mutate(min_dtg_date = min(dtg_recomend_d)) %>%
  mutate(qtr = paste0(substring(year(recart_d),3,4),"/0",quarter(recart_d))) %>%
  mutate(sex = ifelse(sex =="Male", "Men", "Women")) %>%
  mutate(sex = factor(sex))


library(zoo)

tblBAS <- tblBAS %>%
  mutate(recart_d_qtr = as.Date(as.yearqtr(qtr, format = "%y/%q"), frac =1)) %>%
  mutate(recommend_d_qtr = as.Date(as.yearqtr(recommend_qtr, format = "%y/%q"), frac =1)) 

frq(tblBAS, recart_date)

frq(tblBAS, qtr)

frq(tblBAS, recart_d_qtr)

frq(tblBAS, recommend_d_qtr) 

leso <- tblBAS %>% select(country, patient, recart_d, recart_date, recart_date2, dtg_recomend_d, recommend_date, recommend_qtr, min_dtg_date, qtr, sex)  %>%
  filter(country == "LSO")


frq(tblBAS, recart_date)

#frq(tblBAS, ctry)

#frq(tblBAS, age_cat2)

#tblBAS$Datetime <- as.Date(tblBAS$recart_ym)
#DT<- data.table(tblBAS)
#DT[,sum(dtg_first),by = Datetime]

dtg1 <- tblBAS %>% filter(age_cat3 == "[20,30)" | age_cat3 =="[30,40)") %>%
  filter(recart_d <= end_date) %>%
  group_by(ctry, NProg, program, country, sex, qtr, recart_d_qtr, recommend_d_qtr,dtg_recomend_d, recommend_date,  min_dtg_date) %>%
  summarise(
    numer = sum(dtg_first),
    denom = n()) %>%
  mutate(line = "dtg") %>%
  mutate(outcome = "a") %>%
  mutate(per = numer/denom *100) %>%
  mutate(value = round(per, 1)) %>%
  select(-c(per))   


dtg2 <- tblBAS %>% filter(age_cat3 == "[20,30)" | age_cat3 =="[30,40)") %>%
    filter(recart_d <= end_date) %>%
  group_by(ctry, NProg, program, country, sex, qtr, recart_d_qtr, recommend_d_qtr, dtg_recomend_d, recommend_date, min_dtg_date) %>%
  summarise(
    numer_b = sum(dtg_first),
    denom = n(),
    numer = denom - numer_b) %>%
  mutate(line = "dtg") %>%
  mutate(outcome = "b") %>%
  mutate(per = numer/denom *100) %>%
  mutate(value = round(per, 1)) %>%
  select(-c(per, numer_b)) 


data <- rbind(dtg1, dtg2)


data$outcome <- factor(data$outcome, levels = c( "a", "b"),
                       labels = c("DTG ART","Other ART" ))

data$data <- factor(data$sex, levels = c( "Male", "Female"),
                       labels = c("Men","Women" ))


fill <- c( "darkblue", "#56B4E9")

 
## stacked bar chart with legend on side 

data_temp <- data %>% filter(outcome =="DTG ART")
 

plotA_all <- ggplot(data_temp, aes(x=recart_d_qtr, y=value, color=sex)) + 
 geom_line(stat="identity", size= 1.5, linetype = 1, show.legend = TRUE) +
  scale_fill_manual(labels = ~ stringr::str_wrap(.x, width = 40), values = fill)  +
  scale_color_manual(labels = ~ stringr::str_wrap(.x, width = 40), values = fill) +
   theme_bw() + 
  theme(plot.title = element_text(size = 14, hjust = 0.5),
        legend.text = element_text(size = 13),
        legend.title = element_text(size = 13),
        legend.position="bottom", 
        legend.justification = "center",
        legend.box.background = element_blank(),
        legend.box.margin = margin(t = 1, l = 1), 
       #legend.box.background = element_rect(color = "black"),
        legend.direction="vertical", 
        legend.margin=margin(0,0,0,0),
        legend.key.size = unit(0.7, "cm"),
        strip.background = element_blank(),
        strip.text.x = element_text(size=13),
        axis.title.y = element_text(margin = margin(t = 1, r = 1, b = 1, l = 1),size=12),
        axis.title.x = element_text(size=12),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=12, angle=90,vjust = 0.6, hjust = 0.5)) +
  labs(x="", y="Percentage (%)") +
  scale_y_continuous(n.breaks = 5, labels = function(x) paste0(x*1, "%")) +
  #scale_x_date(breaks = "year") +
  ggtitle("(A) Initiating ART with dolutegravir") +
  guides(color = guide_legend(title = "", reverse=FALSE, ncol = 1)) +
  facet_wrap(vars(NProg), nrow=5) +
  geom_vline(data = data_temp, aes(xintercept = recommend_d_qtr),linetype="dashed") 


plotA_all

ggsave("output/IWHOD2024_Poster/Cohorts/DTG_initiation_by_prog_sex_20-39_by_month.png", width = 7, height =8)  

```

```{r  echo=FALSE, fig.height=8, fig.width=6}
plotA_all
```


## South African cohorts only

```{r include=FALSE}
tblBAS <- tblBAS_1 %>%
 #filter(recart_d < "2023-08-01") %>%
  mutate(recart_date2 = floor_date(as_date(recart_d), "month")) %>%
  mutate(recart_date = ceiling_date(as_date(recart_d), "month")) %>%
  mutate(recommend_date = floor_date(as_date(dtg_recomend_d), "month")) %>%
  mutate(recommend_qtr = paste0(substring(year(dtg_recomend_d),3,4),"/0",quarter(dtg_recomend_d))) %>%
  mutate(min_dtg_date = min(dtg_recomend_d)) %>%
  mutate(qtr = paste0(substring(year(recart_d),3,4),"/0",quarter(recart_d))) %>%
  mutate(sex = ifelse(sex =="Male", "Men", "Women")) %>%
  mutate(sex = factor(sex)) %>%
  filter(country == "ZAF")


library(zoo)

tblBAS <- tblBAS %>%
  mutate(recart_d_qtr = as.Date(as.yearqtr(qtr, format = "%y/%q"), frac =1)) %>%
  mutate(recommend_d_qtr = as.Date(as.yearqtr(recommend_qtr, format = "%y/%q"), frac =1)) 

frq(tblBAS, recart_date)

frq(tblBAS, qtr)

frq(tblBAS, recart_d_qtr)

frq(tblBAS, recommend_d_qtr) 

leso <- tblBAS %>% select(country, patient, recart_d, recart_date, recart_date2, dtg_recomend_d, recommend_date, recommend_qtr, min_dtg_date, qtr, sex)  %>%
  filter(country == "LSO")


frq(tblBAS, recart_date)

#frq(tblBAS, ctry)

#frq(tblBAS, age_cat2)

#tblBAS$Datetime <- as.Date(tblBAS$recart_ym)
#DT<- data.table(tblBAS)
#DT[,sum(dtg_first),by = Datetime]

dtg1 <- tblBAS %>% filter(age_cat3 == "[20,30)" | age_cat3 =="[30,40)") %>%
  filter(recart_d <= end_date) %>%
  group_by(ctry, NProg, program, country, sex, qtr, recart_d_qtr, recommend_d_qtr,dtg_recomend_d, recommend_date,  min_dtg_date) %>%
  summarise(
    numer = sum(dtg_first),
    denom = n()) %>%
  mutate(line = "dtg") %>%
  mutate(outcome = "a") %>%
  mutate(per = numer/denom *100) %>%
  mutate(value = round(per, 1)) %>%
  select(-c(per))   


dtg2 <- tblBAS %>% filter(age_cat3 == "[20,30)" | age_cat3 =="[30,40)") %>%
    filter(recart_d <= end_date) %>%
  group_by(ctry, NProg, program, country, sex, qtr, recart_d_qtr, recommend_d_qtr, dtg_recomend_d, recommend_date, min_dtg_date) %>%
  summarise(
    numer_b = sum(dtg_first),
    denom = n(),
    numer = denom - numer_b) %>%
  mutate(line = "dtg") %>%
  mutate(outcome = "b") %>%
  mutate(per = numer/denom *100) %>%
  mutate(value = round(per, 1)) %>%
  select(-c(per, numer_b)) 


data <- rbind(dtg1, dtg2)


data$outcome <- factor(data$outcome, levels = c( "a", "b"),
                       labels = c("DTG ART","Other ART" ))

data$data <- factor(data$sex, levels = c( "Male", "Female"),
                       labels = c("Men","Women" ))


fill <- c( "darkblue", "#56B4E9")

 
## stacked bar chart with legend on side 

data_temp <- data %>% filter(outcome =="DTG ART")
 

plotA_all <- ggplot(data_temp, aes(x=recart_d_qtr, y=value, color=sex)) + 
 geom_line(stat="identity", size= 1.5, linetype = 1, show.legend = TRUE) +
  scale_fill_manual(labels = ~ stringr::str_wrap(.x, width = 40), values = fill)  +
  scale_color_manual(labels = ~ stringr::str_wrap(.x, width = 40), values = fill) +
   theme_bw() + 
  theme(plot.title = element_text(size = 14, hjust = 0, vjust = 0),
        legend.text = element_text(size = 13),
        legend.title = element_text(size = 13),
        legend.position="bottom", 
        legend.justification = "center",
        legend.box.background = element_blank(),
        legend.box.margin = margin(t = 1, l = 1), 
       #legend.box.background = element_rect(color = "black"),
        legend.direction="vertical", 
        legend.margin=margin(0,0,0,0),
        legend.key.size = unit(0.7, "cm"),
        strip.background = element_blank(),
        strip.text.x = element_text(size=13),
        axis.title.y = element_text(margin = margin(t = 1, r = 1, b = 1, l = 1),size=12),
        axis.title.x = element_text(size=12),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=12, angle=90,vjust = 0.6, hjust = 0.5)) +
  labs(x="", y="Percentage (%)") +
  scale_y_continuous(n.breaks = 5, labels = function(x) paste0(x*1, "%")) +
  #scale_x_date(breaks = "year") +
  ggtitle("(A) Initiating ART with dolutegravir") +
  guides(color = guide_legend(title = "", reverse=FALSE, ncol = 1)) +
  facet_wrap(vars(NProg), nrow=1) +
  geom_vline(data = data_temp, aes(xintercept = recommend_d_qtr),linetype="dashed") 


plotA_all

ggsave("output/IWHOD2024_Poster/Cohorts/DTG_initiation_by_prog_sex_20-39_by_month_ZAF.png", width = 6.5, height =4)  

```

```{r  echo=FALSE, fig.height=5, fig.width=5}
plotA_all
```



<!-- ----------------------------------------------------- -->

# Computing Environment

```{r echo=FALSE, results='asis'}
report(sessionInfo())
```