---
title: "SA256 DTG uptake <br> Manuscript_Table1_baseline_chara"
subtitle: " "
author: "Elizabeth Zaniewski, Eliane Rohner"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    code_folding: show
    self_contained: true
    highlight: pygments
    keep_md: no
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
---

```{r r-setup, include = FALSE}
options(scipen = 999)
options(max.print = "300")
set.seed(12345)

library(pacman)
p_load(sf, tidyverse, kableExtra, report, magrittr, fst, data.table, survival, lubridate, ggsurvfit, gtsummary, tidycmprsk, Hmisc, survminer,scales, survMisc, psych, readxl, sjmisc, flextable)

p_load(
  kableExtra,
  scales, ggplot2, dplyr,
  fst, data.table, sjmisc,
  osfr
)




#install.packages("finalfit")

library(survival)
library(lubridate)
library(ggsurvfit)
library(gtsummary)
library(tidycmprsk)
library(Hmisc)
library(survminer)
library(forcats)
library(dplyr)
#library(condsurv)
library(patchwork)
library(ggplot2)
library(cowplot)
library(lmtest)
library(cowplot)
library(tidyverse)
library(sjmisc)
library("forestplot")
library("tiff")

library("tidyverse") 
library(arsenal)

# Analysis
library("MASS")
library("logistf")
library("car")
library("lme4")
library("marginaleffects")

# Tables
library("gtsummary")
library("eeptools")
library("kableExtra")


library(table1)
library(gtsummary)
library(stats)
library(rstatix)


import::from("sjmisc", "frq")
import::from("gmodels", "CrossTable")
import::from("psych", "geometric.mean")

```

```{r conflicts, include = FALSE}
tidyverse::tidyverse_conflicts()

conflicted::conflict_scout()
```

```{r knit-setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

knitr::opts_chunk$set(cache = FALSE,
                      prompt = FALSE,
                      tidy = FALSE,
                      comment = NA,
                      message = FALSE,
                      warning = FALSE,
                      echo = FALSE)

knitr::opts_knit$set(width = 200)
```

<!-- ----------------------------------------------------- -->

```{r}
#Clear existing data and graphics
rm(list=ls())
graphics.off()
```

```{r}

#####################################################################
# to adjust decimal places of categorical variables

my.render.cat <- function(x) {
  c("", sapply(stats.default(x), function(y) with(y,
                                                  sprintf("%d (%0.0f%%)", FREQ, PCT))))
}

#####################################################################

```


```{r}
my.render.continuous <- function(x) { 
  c("", "Sum" = format(sum(x), big.mark = ","))
}
```


```{r}
render.median.IQR <- function(x, ...) {
  c('', 
    `Mean (SD)` = sprintf("%s (%s)", round(mean(x), 2), round(sd(x), 2)),
    `Median [IQR]` = sprintf("%s [%s, %s]", median(x), 
                             quantile(x, 0.25), quantile(x, 0.75)))
}
```


<!-- ----------------------------------------------------- -->


# National DTG adoption dates

Pulled by @aezaniewski from different sources.  

```{r include=FALSE}
dtg_recomend = data.table(read_fst("data_temp/dtg_recomend_v001.fst"))

dtg_recomend$ctry <- factor(dtg_recomend$country, levels = c( "LSO","MWI", "MOZ" ,"ZAF", "ZMB", "ZWE"),
 labels = c("Lesotho",  "Malawi", "Mozambique", "South Africa", "Zambia", "Zimbabwe"))



```

```{r echo=FALSE}
dtg_recomend %>% select(country, ctry, dtg_recomend_d_initiate, dtg_recomend_d) %>%
  kbl() %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  column_spec(1, italic = TRUE) %>%
  row_spec(1, italic = FALSE)
```

<!-- ----------------------------------------------------- -->



**Trends in dolutegravir uptake by sex in Southern Africa**


<!-- ----------------------------------------------------- -->


# Load wrangled data: Initiating


```{r include=FALSE}
tblBAS_1 = data.table(read_fst("data_temp/tblBAS_cohort1_LTFU_WHO_v010.fst"))

tblBAS_1$ctry <- factor(tblBAS_1$country, levels = c( "LSO","MWI", "MOZ" ,"ZAF", "ZMB", "ZWE"),
 labels = c("Lesotho",  "Malawi", "Mozambique", "South Africa", "Zambia", "Zimbabwe"))


```


Remove non-public program
```{r}
tblBAS_1 <- tblBAS_1 %>% filter(!program =="NEWLANDS") 

```


Create table & variable for total number of patients per country

```{r}
frq(tblBAS_1, ctry)
```

```{r}
my_comma <- scales::label_comma(accuracy = 1, big.mark = ",")

PatperCtry_1 <- tblBAS_1 %>% 
  group_by(ctry) %>%
  summarise("NperCtry"=n()) %>%
  mutate(NperCtry2 = my_comma(NperCtry)) %>%
  mutate(text1 = " (n=") %>%
  mutate(text2 = ")") %>%
  mutate(NCtry = paste(ctry, "\n", text1, NperCtry2, text2, sep= "")) %>%
  select(ctry, NCtry) 

tblBAS_1 <- merge(tblBAS_1, PatperCtry_1, by = "ctry", 
           all.x = TRUE, all.y = FALSE) 

```




```{r include=FALSE}

# Create sex and age categories
tblBAS_1 <- tblBAS_1 %>%
  mutate(sex_age2_txt = ifelse(sex_age2 =="Male [20,50)", "Male 20-49",
                               ifelse(sex_age2 =="Female [20,50)", "Female 20-49",
                                    ifelse(sex_age2 =="Male [50,100]", "Male 50+",
                                         ifelse(sex_age2 =="Female [50,100]", "Female 50+", "other"
                                      )))))

frq(tblBAS_1, sex_age2)
frq(tblBAS_1, sex_age2_txt)

tblBAS_1 <- tblBAS_1 %>%
  mutate(sex_age3_txt = ifelse(sex_age3 =="Male.[20,30)", "Male 20-29",
                               ifelse(sex_age3 =="Female.[20,30)", "Female 20-29",
                                ifelse(sex_age3 =="Male.[30,40)", "Male 30-39",
                               ifelse(sex_age3 =="Female.[30,40)", "Female 30-39",
                              ifelse(sex_age3 =="Male.[40,50)", "Male 40-49",
                               ifelse(sex_age3 =="Female.[40,50)", "Female 40-49",
                                ifelse(sex_age3 =="Male.[50,100]", "Male 50+",
                                ifelse(sex_age3 =="Female.[50,100]", "Female 50+", "other"
                                      )))))))))

```



```{r echo=FALSE}
table1::label(tblBAS_1$sex) <- "Sex"
table1::label(tblBAS_1$age_cat3) <- "Age group"
table1::label(tblBAS_1$dtg_first_txt) <- "ART at initiation"
table1::label(tblBAS_1$sex_age2) <- "Sex & age group"
table1::label(tblBAS_1$sex_age3) <- "Sex & age group"
table1::label(tblBAS_1$program) <- "Program"
table1::label(tblBAS_1$ctry) <- "Country"
table1::label(tblBAS_1$sex_age2_txt) <- "Sex & age group"
table1::label(tblBAS_1$sex_age3_txt) <- "Sex & age group"

```

<!-- ----------------------------------------------------- -->


# Load wrangled data: Switching

```{r include=FALSE}
tblBAS_2 = data.table(read_fst("data_temp/tblBAS_cohort2_LTFU_WHO_v011.fst")) 

```


Remove non-public program
```{r}
tblBAS_2 <- tblBAS_2 %>% filter(!program =="NEWLANDS") 

```



Create table & variable for total number of patients per country


```{r}
frq(tblBAS_2, ctry)
```

```{r}
my_comma <- scales::label_comma(accuracy = 1, big.mark = ",")

PatperCtry_2 <- tblBAS_2 %>% 
  group_by(ctry) %>%
  summarise("NperCtry"=n()) %>%
  mutate(NperCtry2 = my_comma(NperCtry)) %>%
  mutate(text1 = " (n=") %>%
  mutate(text2 = ")") %>%
  mutate(NCtry = paste(ctry, "\n", text1, NperCtry2, text2, sep= "")) %>%
  select(ctry, NCtry) 

tblBAS_2 <- merge(tblBAS_2, PatperCtry_2, by = "ctry", 
           all.x = TRUE, all.y = FALSE) 

```





```{r}
                                     

# Create sex and age categories
tblBAS_2 <- tblBAS_2 %>%
  mutate(sex_age2_txt = ifelse(sex_age2 =="Male [20,50)", "Male 20-49",
                               ifelse(sex_age2 =="Female [20,50)", "Female 20-49",
                                    ifelse(sex_age2 =="Male [50,100]", "Male 50+",
                                         ifelse(sex_age2 =="Female [50,100]", "Female 50+", "other"
                                      )))))

frq(tblBAS_2, sex_age2)
frq(tblBAS_2, sex_age2_txt)

tblBAS_2 <- tblBAS_2 %>%
  mutate(sex_age3_txt = ifelse(sex_age3 =="Male.[20,30)", "Male 20-29",
                               ifelse(sex_age3 =="Female.[20,30)", "Female 20-29",
                                ifelse(sex_age3 =="Male.[30,40)", "Male 30-39",
                               ifelse(sex_age3 =="Female.[30,40)", "Female 30-39",
                              ifelse(sex_age3 =="Male.[40,50)", "Male 40-49",
                               ifelse(sex_age3 =="Female.[40,50)", "Female 40-49",
                                ifelse(sex_age3 =="Male.[50,100]", "Male 50+",
                                ifelse(sex_age3 =="Female.[50,100]", "Female 50+", "other"
                                      )))))))))
```




```{r}
table1::label(tblBAS_2$sex) <- "Sex"
table1::label(tblBAS_2$age_cat3) <- "Age group"
table1::label(tblBAS_2$dtg_switch_txt) <- "DTG switch"
table1::label(tblBAS_2$sex_age2) <- "Sex & age group"
table1::label(tblBAS_2$sex_age3) <- "Sex & age group"
table1::label(tblBAS_2$program) <- "Program"
table1::label(tblBAS_2$ctry) <- "Country"
table1::label(tblBAS_2$sex_age2_txt) <- "Sex & age group"
table1::label(tblBAS_2$sex_age3_txt) <- "Sex & age group"
```

<!-- ----------------------------------------------------- -->

# Initiation cohort

```{r}

frq(tblBAS_1, sex)

temp1 <- tblBAS_1 %>% select(country, program, dtg_recomend_d, max_close_d) %>%
  filter(!program =="NEWLANDS") %>%
  arrange(dtg_recomend_d, country, program) %>%
  group_by(country, program, dtg_recomend_d, max_close_d) %>%
  mutate(Npatients = n()) %>%
  distinct()
  
temp1 %>% 
  kbl() %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  column_spec(1, italic = TRUE) %>%
  row_spec(1, italic = FALSE)

```


## Overview by country

```{r}
sjPlot::tab_xtab(tblBAS_1$ctry,
                 tblBAS_1$sex,
                 show.row.prc = TRUE, show.summary = FALSE)
```


```{r}
sjPlot::tab_xtab(tblBAS_1$ctry,
                 tblBAS_1$sex_age3_txt,
                 show.row.prc = TRUE, show.summary = FALSE)
```

## Overview by program

```{r}
sjPlot::tab_xtab(tblBAS_1$program,
                 tblBAS_1$sex,
                 show.row.prc = TRUE, show.summary = FALSE)
```


## Initiating "Other ART" vs "DTG ART".

by country

```{r}
sjPlot::tab_xtab(tblBAS_1$ctry,
                 tblBAS_1$dtg_first_txt,
                 show.row.prc = TRUE, show.summary = FALSE)
```

by sex & age

```{r}
sjPlot::tab_xtab(tblBAS_1$sex_age3_txt,
                 tblBAS_1$dtg_first_txt,
                 show.row.prc = TRUE, show.summary = FALSE)
```


<!-- ----------------------------------------------------- -->

# Switching cohort

```{r}
frq(tblBAS_2, sex)


temp2 <- tblBAS_2 %>% select(country, program, dtg_recomend_d, max_close_d) %>%
   filter(!program =="NEWLANDS") %>%
  arrange(dtg_recomend_d, country, program) %>%
  group_by(country, program, dtg_recomend_d, max_close_d) %>%
  mutate(Npatients = n()) %>%
  distinct()
  
temp2 %>% 
  kbl() %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  column_spec(1, italic = TRUE) %>%
  row_spec(1, italic = FALSE)
```


```{r echo=FALSE}
tblBAS_2 %>%
  ggplot(aes(x = rev_outcome)) +
  geom_bar() +
  theme_minimal() +
  xlab("Outcome") +
  ylab("Number of patients") +
  facet_wrap(vars(program), scales = "free_x") +
  coord_flip()
```



## Overview by country

```{r}
sjPlot::tab_xtab(tblBAS_2$ctry,
                 tblBAS_2$sex,
                 show.row.prc = TRUE, show.summary = FALSE)
```


```{r}
sjPlot::tab_xtab(tblBAS_2$ctry,
                 tblBAS_2$sex_age3_txt,
                 show.row.prc = TRUE, show.summary = FALSE)
```


## Overview by program

```{r}
sjPlot::tab_xtab(tblBAS_2$program,
                 tblBAS_2$sex,
                 show.row.prc = TRUE, show.summary = FALSE)
```


## Median time on ART


```{r}

tblBAS_2b <- tblBAS_2 %>% mutate(timeonART_d = as.numeric(dtg_recomend_d - recart_d)) %>%
  mutate(timeonART_m = round(timeonART_d/30.5, 1)) 
  

timeonART_ctry_sex <- tblBAS_2b %>% 
  group_by(ctry, sex) %>% 
   filter(!is.na(timeonART_m)) %>%  
   summarise(#"quantile10"= round(quantile(timeonART_m,c(.10)),digits=1),
            "Median"= round(quantile(timeonART_m,c(.50)),digits=1),
            "quantile25"= round(quantile(timeonART_m,c(.25)),digits=1),
            "quantile75"= round(quantile(timeonART_m,c(.75)),digits=1),
            #"quantile90"= round(quantile(timeonART_m,c(.90)),digits=1),
            #"Mean" = round(mean(timeonART_m),digits=1)
            )


timeonART_sex <- tblBAS_2b %>% 
  group_by(sex) %>% 
   filter(!is.na(timeonART_m)) %>%  
  summarise(#"quantile10"= round(quantile(timeonART_m,c(.10)),digits=1),
            "Median"= round(quantile(timeonART_m,c(.50)),digits=1),
            "quantile25"= round(quantile(timeonART_m,c(.25)),digits=1),
            "quantile75"= round(quantile(timeonART_m,c(.75)),digits=1),
            #"quantile90"= round(quantile(timeonART_m,c(.90)),digits=1),
            #"Mean" = round(mean(timeonART_m),digits=1)
            )
  

```



```{r}
timeonART_sex2 <- timeonART_sex

writeLines("td, th { padding : 6px } th { background-color : brown ; color : white; border : 1px solid white; } td { color : brown ; border : 1px solid brown }", con = "mystyle.css")
dset1 <- head(timeonART_sex2)
knitr::kable(dset1, format = "html")

```


```{r}
timeonART_ctry_sex2 <- timeonART_ctry_sex

writeLines("td, th { padding : 6px } th { background-color : brown ; color : white; border : 1px solid white; } td { color : brown ; border : 1px solid brown }", con = "mystyle.css")
dset1 <- head(timeonART_ctry_sex2)
knitr::kable(dset1, format = "html")

```


```{r eval=FALSE, include=FALSE}

table1::table1(~ Median + quantile25 + quantile75 | sex , data = timeonART_sex, overall=c(left="Total"), render.categorical = my.render.cat)

```



```{r}
rm(timeonART_ctry_sex2)
rm(timeonART_sex2)
```



## Switching to DTG

Number of adults on ART switching to "DTG ART".

by country

```{r}
sjPlot::tab_xtab(tblBAS_2$ctry,
                 tblBAS_2$dtg_switch_txt,
                 show.row.prc = TRUE, show.summary = FALSE)
```


by sex & age

```{r}
sjPlot::tab_xtab(tblBAS_2$sex_age3_txt,
                 tblBAS_2$dtg_switch_txt,
                 show.row.prc = TRUE, show.summary = FALSE)
```


```{r}

sjPlot::tab_xtab(tblBAS_2$ctry,
                 tblBAS_2$dtg_switch_txt,
                 show.row.prc = TRUE, show.summary = FALSE)
```



```{r}

sjPlot::tab_xtab(tblBAS_2$ctry,
                 tblBAS_2$rev_outcome,
                 show.row.prc = TRUE, show.summary = FALSE)
```



```{r eval=FALSE, include=FALSE}
test <- tblBAS_2 %>% filter(country =="ZMB") %>% select(patient, country, rev_outcome, rev_outcome_d, max_close_d)
```


```{r}
FU_time <- tblBAS_2 %>% select(patient, country, NCtry, ctry, program, sex, 
                    age, age_cat1, age_cat2,  age_cat3, 
                    sex_age1, sex_age2, sex_age2_txt, sex_age3, sex_age3_txt,
                    start_fu, end_fu, FU_day, time,
                    dtg_switch, dtg_switch_txt, 
                    max_close_d, dtg_switch_txt, dtg_status_outcome, dtg_status_outcome_txt, 
                     dtg_recomend_d, dtg_recomend_t, 
                    time_t, time_t2) 

```



```{r include=FALSE}

#https://argoshare.is.ed.ac.uk/healthyr_book/recode-the-data-1.html

##########################################
# Recode

FU_time <- FU_time %>% mutate(dtg_comp = ifelse(dtg_switch ==0 & dtg_status_outcome_txt != "RIC", 1, 0))

FU_time <- FU_time %>% mutate(dtg_comp_b = ifelse(dtg_switch ==0 & dtg_status_outcome_txt == "RIC", 0, 1))


frq(FU_time, dtg_status_outcome)
frq(FU_time, dtg_status_outcome_txt)
frq(FU_time, dtg_comp)
frq(FU_time, dtg_comp_b)

FU_time = as.data.table(FU_time)

frq(FU_time, dtg_status_outcome)
frq(FU_time, dtg_status_outcome_txt)
frq(FU_time, dtg_switch)  # 1 == started DTG, 0 == didn't start DTG
frq(FU_time, dtg_comp)  # 1 == competing risk; prevented pat from being at risk to start dtg 

zmb <- FU_time %>% filter(country =="ZMB")
  frq(zmb,dtg_status_outcome)

FU_time <- FU_time %>% 
  mutate(
    
  # Overall survival (alive ==0,  dtg ==1)
  dtg_switch_txt_os = case_when(
    dtg_switch == 1 ~ 1,  #started DTG
    dtg_switch == 0 ~ 0), #didn't start DTG

  # Diease-specific survival (alive ==0, dtg ==1)
  dtg_switch_txt_dss = case_when(
    dtg_status_outcome == 4 ~ 0, #still alive not on dtg
    dtg_switch == 1 ~ 1,         #started DTG
    dtg_comp == 1 ~ 0),      # censored
  
   # Competing risks regression (alive ==0, dtg ==1, death ==2, ltfu ==3, trans ==4)
  dtg_switch_txt_crr = case_when(
    dtg_status_outcome == 4 ~ 0, # still alive RIC no switch to DTG
    dtg_status_outcome == 0 ~ 1, # switched to DTG                 
    dtg_status_outcome == 1 ~ 2, # had other competing risk outcome Death
    dtg_status_outcome == 2 ~ 3, # had other competing risk outcome LTFU
    dtg_status_outcome == 3 ~ 4), # had other competing risk outcome Transfer
  
     # Competing risks regression (alive ==0, dtg ==1, death ==2, ltfu ==3, trans ==4)
  dtg_switch_txt_crr1b = case_when(
    dtg_status_outcome == 4 ~ 0, # still alive RIC no switch to DTG
    dtg_status_outcome == 0 ~ 4, # switched to DTG                 
    dtg_status_outcome == 1 ~ 1, # had other competing risk outcome Death
    dtg_status_outcome == 2 ~ 3, # had other competing risk outcome LTFU
    dtg_status_outcome == 3 ~ 2), # had other competing risk outcome Transfer

 # Competing risks regression (alive ==0, dtg ==1, all competing risks ==2)
  dtg_switch_txt_crr2 = case_when( 
    dtg_status_outcome == 4 ~ 0, # still alive RIC no switch to DTG
    dtg_status_outcome == 0 ~ 1, # switched to DTG                 
    dtg_status_outcome == 1 ~ 2, # had other competing risk outcome Death
    dtg_status_outcome == 2 ~ 2, # had other competing risk outcome LTFU
    dtg_status_outcome == 3 ~ 2), # had other competing risk outcome Transfer
   

 # Competing risks regression (alive ==0, dtg ==1, all competing risks ==0)
  dtg_switch_txt_crr3 = case_when( 
    dtg_status_outcome == 4 ~ 0, # still alive RIC no switch to DTG
    dtg_status_outcome == 0 ~ 1, # switched to DTG                 
    dtg_status_outcome == 1 ~ 0, # had other competing risk outcome Death
    dtg_status_outcome == 2 ~ 0, # had other competing risk outcome LTFU
    dtg_status_outcome == 3 ~ 0), # had other competing risk outcome Transfer
   ) 


# factor variables
FU_time$dtg_switch_txt_crr.factor <- factor(FU_time$dtg_switch_txt_crr, 
levels=c("0", "1", "2", "3", "4"),
labels = c("No DTG", "DTG switch", "Death", "LTFU", "Transfer"))

FU_time$dtg_switch_txt_crr1b.factor <- factor(FU_time$dtg_switch_txt_crr1b, 
levels=c("0", "1", "2", "3", "4"),
labels = c("a", "b", "c", "d", "e"))

# a=RIC, no dtg, b=death, c=transfer, d=LTFU, e=switched to dtg

FU_time$dtg_switch_txt_crr2.factor <- factor(FU_time$dtg_switch_txt_crr2,
levels=c("0", "1", "2"),
labels = c("a", "c", "b"))

FU_time$dtg_switch_txt_crr2b.factor <- factor(FU_time$dtg_switch_txt_crr2,
levels=c("0", "1", "2"),
labels = c("RIC", "DTG switch", "competing risks"))

FU_time$dtg_switch_txt_crr3b.factor <- factor(FU_time$dtg_switch_txt_crr3,
levels=c("0", "1"),
labels = c("No Switch", "DTG switch"))



frq(FU_time, dtg_switch_txt_os)
frq(FU_time, dtg_switch_txt_dss)
frq(FU_time, dtg_switch_txt_crr)
frq(FU_time, dtg_switch_txt_crr.factor)
frq(FU_time, dtg_switch_txt_crr2)
frq(FU_time, dtg_switch_txt_crr2.factor)
frq(FU_time, dtg_switch_txt_crr2b.factor)
frq(FU_time, dtg_switch_txt_crr3)
frq(FU_time, dtg_switch_txt_crr3b.factor)


zmb <- FU_time %>% filter(country =="ZMB")
frq(zmb,dtg_status_outcome)
  frq(zmb,dtg_status_outcome_txt)
  frq(zmb,dtg_switch_txt_crr1b)
   frq(zmb,dtg_switch_txt_crr1b.factor)
   frq(zmb,dtg_switch_txt_crr2)
    frq(zmb,dtg_switch_txt_crr2b.factor)
   
```

```{r}

# Label and recode other variables
FU_time <- FU_time %>%
 mutate(sex2 = factor(sex) %>% 
 fct_recode("1" = "Male",
            "0" = "Female")) %>%
  mutate(dtg.factor = factor(dtg_switch)) %>%
  mutate(sex3 = ifelse(sex =="Male", "Men", "Women")) %>%
  mutate(sex3.factor = factor(sex3))
 

```


<!-- ----------------------------------------------------- -->

# Table 1: Baseline characteristics

## Re-wrangle data



Country
```{r}
frq(tblBAS_1, ctry)
```

Sex
```{r}
frq(tblBAS_1, sex)


tblBAS_1$sex.factor = factor(tblBAS_1$sex, 
levels = c('Female', 'Male'), 
labels = c('Women', 'Men'))

frq(tblBAS_1, sex.factor)
```


Sex_age2_txt
```{r}
frq(tblBAS_1, sex_age2_txt)

tblBAS_1$sex_age2.factor = factor(tblBAS_1$sex_age2, 
levels = c('Female [20,50)', 'Female [50,100]', 'Male [20,50)', 'Male [50,100]'), 
labels = c('Women 20-49', 'Women 50+', 'Men 20-49', 'Men 50+'))

frq(tblBAS_1, sex_age2.factor)
```



Labels of variables used in Table 1
```{r}

label(tblBAS_1$ctry) <- "Country"
label(tblBAS_1$sex.factor) <- "Sex"
label(tblBAS_1$sex_age2.factor) <- "Sex and age in years"


```


```{r eval=FALSE, include=FALSE}
#remove column
tblBAS_1[,patients:=NULL]
```


Create dummy variable
```{r}
tblBAS_1[, patients := "Total"]
label(tblBAS_1$patients) <- "Total patients"

```


## Create Table 1: version 1


```{r}
my.render.cat <- function(x) {
  c("", sapply(stats.default(x), function(y) with(y,
                                                  sprintf("%d (%0.0f%%)", FREQ, PCT))))
}

```


```{r}
my.render.continuous <- function(x) { 
  c("", "Sum" = format(sum(x), big.mark = ","))
}
```


```{r}
render.median.IQR_F <- function(x, ...) {
  c('', 
        `Median (IQR) age of women in years` = sprintf("%s (%s to %s)", median(x), 
                             quantile(x, 0.25), quantile(x, 0.75)))
}
```


```{r}
render.median.IQR_M <- function(x, ...) {
  c('', 
        `Median (IQR) age of men in years` = sprintf("%s (%s to %s)", median(x), 
                             quantile(x, 0.25), quantile(x, 0.75)))
}
```


Number of patients
```{r}

mytable1_pat <- table1::table1(~  patients + sex.factor | ctry, overall=c(right="Total"), data = tblBAS_1,  render.continuous = my.render.continuous, render.categorical = my.render.cat)

mytable1_1pat<- as.data.frame(mytable1_pat)

```

Median age females
```{r}

mytable1_ageF <- table1::table1(~  age | ctry, overall=c(right="Total"), data = tblBAS_1[sex == "Female", ],  render.continuous = render.median.IQR_F, render.categorical = my.render.cat)

mytable1_2ageF <- as.data.frame(mytable1_ageF)
```

Median age males
```{r}
mytable1_ageM <- table1::table1(~  age | ctry, overall=c(right="Total"), data = tblBAS_1[sex == "Male", ],  render.continuous = render.median.IQR_M, render.categorical = my.render.cat)

mytable1_3ageM <- as.data.frame(mytable1_ageM)
```

Number by sex and age
```{r}
mytable1_sexage <- table1::table1(~  sex_age2.factor | ctry, overall=c(right="Total"), data = tblBAS_1,  render.continuous = my.render.continuous, render.categorical = my.render.cat)

mytable1_4sexage<- as.data.frame(mytable1_sexage)
```

Merge tables

```{r}
mytable1_1pat_b <- mytable1_1pat[-c(1,2,4),]
mytable1_1pat_b[,1] <- c("Total","  Women", "  Men") 

mytable1_2ageF_b <- mytable1_2ageF[-c(1),]
mytable1_2ageF_b[,1] <- c("Median (IQR) age in years","  Women") 

mytable1_3ageM_b <- mytable1_3ageM[-c(1,2),]
mytable1_3ageM_b[,1] <- c("  Men") 


mytable1_4sexage_b <- mytable1_4sexage[-c(1),]

table1_final1 <- rbind(mytable1_1pat_b, mytable1_2ageF_b, mytable1_3ageM_b, mytable1_4sexage_b)
```



```{r}
rm(list=ls(pattern="mytable"))
```



# Table 2: Baseline characteristics

## Re-wrangle data


Country
```{r}
frq(tblBAS_2, ctry)
```

Sex
```{r}
frq(tblBAS_2, sex)


tblBAS_2$sex.factor = factor(tblBAS_2$sex, 
levels = c('Female', 'Male'), 
labels = c('Women', 'Men'))

frq(tblBAS_2, sex.factor)
```


Sex_age2_txt
```{r}
frq(tblBAS_2, sex_age2_txt)

tblBAS_2$sex_age2.factor = factor(tblBAS_2$sex_age2, 
levels = c('Female [20,50)', 'Female [50,100]', 'Male [20,50)', 'Male [50,100]'), 
labels = c('Women 20-49', 'Women 50+', 'Men 20-49', 'Men 50+'))

frq(tblBAS_2, sex_age2.factor)
```


Labels of variables used in Table 2
```{r}

label(tblBAS_2$ctry) <- "Country"
label(tblBAS_2$sex.factor) <- "Sex"
label(tblBAS_2$sex_age2.factor) <- "Sex and age in years"


```


```{r eval=FALSE, include=FALSE}
#remove column
tblBAS_2[,patients:=NULL]
```


Create dummy variable
```{r}
tblBAS_2[, patients := "Total"]
label(tblBAS_2$patients) <- "Total patients"

```



```{r}
my.render.cat <- function(x) {
  c("", sapply(stats.default(x), function(y) with(y,
                                                  sprintf("%d (%0.0f%%)", FREQ, PCT))))
}

```


```{r}
my.render.continuous <- function(x) { 
  c("", "Sum" = format(sum(x), big.mark = ","))
}
```


```{r}
render.median.IQR_F <- function(x, ...) { 
     #stats.apply.rounding(stats.default(round(x, 1)), digits=3)

  c('', 
        `Median (IQR) women` = sprintf("%s (%s to %s)", median(x), 
                             quantile(x, 0.25), quantile(x, 0.75)))
  
}
```


```{r}
render.median.IQR_M <- function(x, ...) {
    # stats.apply.rounding(stats.default(round(x, 1)), digits=3)
       c('', 
        `Median (IQR) men` = sprintf("%s (%s to %s)", median(x), 
                             quantile(x, 0.25), quantile(x, 0.75)))
     }
```


Number of patients
```{r}

mytable2_pat <- table1::table1(~  patients + sex.factor | ctry, overall=c(right="Total"), data = tblBAS_2,  render.continuous = my.render.continuous, render.categorical = my.render.cat)

mytable2_1pat<- as.data.frame(mytable2_pat)

```

Median age females
```{r}

mytable2_ageF <- table1::table1(~  age | ctry, overall=c(right="Total"), data = tblBAS_2[sex == "Female", ],  render.continuous = render.median.IQR_F, render.categorical = my.render.cat)

mytable2_2ageF <- as.data.frame(mytable2_ageF)
```

Median age males
```{r}
mytable2_ageM <- table1::table1(~  age | ctry, overall=c(right="Total"), data = tblBAS_2[sex == "Male", ],  render.continuous = render.median.IQR_M, render.categorical = my.render.cat)

mytable2_3ageM <- as.data.frame(mytable2_ageM)
```

Number by sex and age
```{r}
mytable2_sexage <- table1::table1(~  sex_age2.factor | ctry, overall=c(right="Total"), data = tblBAS_2,  render.continuous = my.render.continuous, render.categorical = my.render.cat)

mytable2_4sexage<- as.data.frame(mytable2_sexage)
```

Time on ART females
```{r}
mytable2_timeF <- table1::table1(~  timeonART_m | ctry, overall=c(right="Total"), data = tblBAS_2b[sex == "Female", ],  render.continuous = render.median.IQR_F, render.categorical = my.render.cat)

mytable2_5timeF <- as.data.frame(mytable2_timeF)
```


Time on ART males
```{r}
mytable2_timeM <- table1::table1(~  timeonART_m | ctry, overall=c(right="Total"), data = tblBAS_2b[sex == "Male", ],  render.continuous = render.median.IQR_M, render.categorical = my.render.cat)

mytable2_6timeM <- as.data.frame(mytable2_timeM)
```


Merge tables

```{r}
mytable2_1pat_b <- mytable2_1pat[-c(1,2,4),]
mytable2_1pat_b[,1] <- c("Total","  Women", "  Men") 

mytable2_2ageF_b <- mytable2_2ageF[-c(1),]
mytable2_2ageF_b[,1] <- c("Median (IQR) age in years","  Women") 

mytable2_3ageM_b <- mytable2_3ageM[-c(1,2),]
mytable2_3ageM_b[,1] <- c("  Men") 

mytable2_4sexage_b <- mytable2_4sexage[-c(1),]

mytable2_5timeF_b <- mytable2_5timeF[-c(1),]
mytable2_5timeF_b[,1] <- c("Median (IQR) time on ART in months","  Women") 

mytable2_6timeM_b <- mytable2_6timeM[-c(1,2),]
mytable2_6timeM_b[,1] <- c("  Men") 



table2_final1 <- rbind(mytable2_1pat_b, mytable2_2ageF_b, mytable2_3ageM_b, mytable2_4sexage_b,mytable2_5timeF_b, mytable2_6timeM_b)
```


```{r}
rm(list=ls(pattern="mytable"))
```


<!-- ----------------------------------------------------- -->

# Computing Environment

```{r echo=FALSE, results='asis'}
report(sessionInfo())
```