---
title: "IeDEA SA256 DTG uptake: 03 Paper Fig 3 KM switching"
subtitle: " "
author: "Elizabeth Zaniewski, Eliane Rohner"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    code_folding: show
    self_contained: true
    highlight: pygments
    keep_md: no
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
---

```{r r-setup, include = FALSE}
options(scipen = 999)
options(max.print = "300")
set.seed(12345)

library(pacman)
p_load(sf, tidyverse, kableExtra, report, magrittr, fst, data.table, survival, lubridate, ggsurvfit, gtsummary, tidycmprsk, Hmisc, survminer, survMisc)

p_load(
  kableExtra,
  scales, ggplot2, dplyr,
  fst, data.table, sjmisc,
  osfr
)

#install.packages("finalfit")

library(survival)
library(lubridate)
library(ggsurvfit)
library(gtsummary)
library(tidycmprsk)
library(Hmisc)
library(survminer)
library(forcats)
library(dplyr)
#library(condsurv)
library(patchwork)



import::from("sjmisc", "frq")
import::from("gmodels", "CrossTable")
```

```{r conflicts, include = FALSE}
tidyverse::tidyverse_conflicts()

conflicted::conflict_scout()
```

```{r knit-setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

knitr::opts_chunk$set(cache = FALSE,
                      prompt = FALSE,
                      tidy = FALSE,
                      comment = NA,
                      message = FALSE,
                      warning = FALSE,
                      echo = FALSE)

knitr::opts_knit$set(width = 200)
```


```{r}

#####################################################################
# to adjust decimal places of categorical variables

my.render.cat <- function(x) {
  c("", sapply(stats.default(x), function(y) with(y,
                                                  sprintf("%d (%0.0f %%)", FREQ, PCT))))
}

#####################################################################

```



<!-- ----------------------------------------------------- -->

```{r}
#Clear existing data and graphics
rm(list=ls())
graphics.off()
```


```{r include=FALSE}
tblBAS_2 = data.table(read_fst("data_temp/tblBAS_transition_v2_LTFU_who.fst"))

```

```{r include=FALSE}
frq(tblBAS_2, sex_age2)

tblBAS_2 <- tblBAS_2 %>%
  mutate(sex_age2_txt = ifelse(sex_age2 =="Male.[20,50)", "Male 20-49",
                               ifelse(sex_age2 =="Female.[20,50)", "Female 20-49",
                                    ifelse(sex_age2 =="Male.[50,100]", "Male 50+",
                                         ifelse(sex_age2 =="Female.[50,100]", "Female 50+", "other"
                                      )))))
frq(tblBAS_2, sex_age2_txt)


frq(tblBAS_2, sex_age3)

tblBAS_2 <- tblBAS_2 %>%
  mutate(sex_age3_txt = ifelse(sex_age3 =="Male.[20,30)", "Male 20-29",
                               ifelse(sex_age3 =="Female.[20,30)", "Female 20-29",
                                ifelse(sex_age3 =="Male.[30,40)", "Male 30-39",
                               ifelse(sex_age3 =="Female.[30,40)", "Female 30-39",
                              ifelse(sex_age3 =="Male.[40,50)", "Male 40-49",
                               ifelse(sex_age3 =="Female.[40,50)", "Female 40-49",
                                ifelse(sex_age3 =="Male.[50,100]", "Male 50+",
                                ifelse(sex_age3 =="Female.[50,100]", "Female 50+", "other"
                                      )))))))))
frq(tblBAS_2, sex_age3_txt)
```


```{r}
table1::label(tblBAS_2$sex) <- "Sex"
table1::label(tblBAS_2$age_cat3) <- "Age group"
table1::label(tblBAS_2$dtg_switch_txt) <- "DTG switch"
table1::label(tblBAS_2$sex_age2) <- "Sex & age group"
table1::label(tblBAS_2$sex_age3) <- "Sex & age group"
table1::label(tblBAS_2$program) <- "Program"
table1::label(tblBAS_2$ctry) <- "Country"
table1::label(tblBAS_2$sex_age2_txt) <- "Sex & age group"
table1::label(tblBAS_2$sex_age3_txt) <- "Sex & age group"
```



```{r}
FU_time <- tblBAS_2 %>% select(patient, country, ctry, program, sex, 
                    age, age_cat1, age_cat2,  age_cat3, 
                    sex_age1, sex_age2, sex_age2_txt, sex_age3, sex_age3_txt,
                    start_fu, end_fu, FU_day, time,
                    dtg_switch, dtg_switch_txt, 
                    max_close_d, dtg_switch_txt, dtg_status_outcome, dtg_status_outcome_txt, 
                     dtg_recomend_d, dtg_recomend_t, 
                    time_t, time_t2,
                    alt_start_fu, alt_end_fu, 
                    alt_start_fu2, alt_end_fu2, 
                    alt_start_fu3, alt_end_fu3)

```



```{r include=FALSE}

#https://argoshare.is.ed.ac.uk/healthyr_book/recode-the-data-1.html

##########################################
# Recode

FU_time <- FU_time %>% mutate(dtg_comp = ifelse(dtg_switch ==0 & dtg_status_outcome_txt != "RIC", 1, 0))

FU_time <- FU_time %>% mutate(dtg_comp_b = ifelse(dtg_switch ==0 & dtg_status_outcome_txt == "RIC", 0, 1))


frq(FU_time, dtg_status_outcome)
frq(FU_time, dtg_status_outcome_txt)
frq(FU_time, dtg_comp)
frq(FU_time, dtg_comp_b)

FU_time = as.data.table(FU_time)

frq(FU_time, dtg_status_outcome)
frq(FU_time, dtg_status_outcome_txt)
frq(FU_time, dtg_switch)  # 1 == started DTG, 0 == didn't start DTG
frq(FU_time, dtg_comp)  # 1 == competing risk; prevented pat from being at risk to start dtg 

FU_time <- FU_time %>% 
  mutate(
    
  # Overall survival (alive ==0,  dtg ==1)
  dtg_switch_txt_os = case_when(
    dtg_switch == 1 ~ 1,  #started DTG
    dtg_switch == 0 ~ 0), #didn't start DTG

  # Diease-specific survival (alive ==0, dtg ==1)
  dtg_switch_txt_dss = case_when(
    dtg_status_outcome == 4 ~ 0, #still alive not on dtg
    dtg_switch == 1 ~ 1,         #started DTG
    dtg_comp == 1 ~ 0),      # censored
  
   # Competing risks regression (alive ==0, dtg ==1, death ==2, ltfu ==3, trans ==4)
  dtg_switch_txt_crr = case_when(
    dtg_status_outcome == 4 ~ 0, # still alive RIC no switch to DTG
    dtg_status_outcome == 0 ~ 1, # switched to DTG                 
    dtg_status_outcome == 1 ~ 2, # had other competing risk outcome Death
    dtg_status_outcome == 2 ~ 3, # had other competing risk outcome LTFU
    dtg_status_outcome == 3 ~ 4), # had other competing risk outcome Transfer
  
     # Competing risks regression (alive ==0, dtg ==1, death ==2, ltfu ==3, trans ==4)
  dtg_switch_txt_crr1b = case_when(
    dtg_status_outcome == 4 ~ 0, # still alive RIC no switch to DTG
    dtg_status_outcome == 0 ~ 4, # switched to DTG                 
    dtg_status_outcome == 1 ~ 1, # had other competing risk outcome Death
    dtg_status_outcome == 2 ~ 3, # had other competing risk outcome LTFU
    dtg_status_outcome == 3 ~ 2), # had other competing risk outcome Transfer

 # Competing risks regression (alive ==0, dtg ==1, all competing risks ==2)
  dtg_switch_txt_crr2 = case_when( 
    dtg_status_outcome == 4 ~ 0, # still alive RIC no switch to DTG
    dtg_status_outcome == 0 ~ 1, # switched to DTG                 
    dtg_status_outcome == 1 ~ 2, # had other competing risk outcome Death
    dtg_status_outcome == 2 ~ 2, # had other competing risk outcome LTFU
    dtg_status_outcome == 3 ~ 2), # had other competing risk outcome Transfer
   

 # Competing risks regression (alive ==0, dtg ==1, all competing risks ==0)
  dtg_switch_txt_crr3 = case_when( 
    dtg_status_outcome == 4 ~ 0, # still alive RIC no switch to DTG
    dtg_status_outcome == 0 ~ 1, # switched to DTG                 
    dtg_status_outcome == 1 ~ 0, # had other competing risk outcome Death
    dtg_status_outcome == 2 ~ 0, # had other competing risk outcome LTFU
    dtg_status_outcome == 3 ~ 0), # had other competing risk outcome Transfer
   ) 



# factor variables
FU_time$dtg_switch_txt_crr.factor <- factor(FU_time$dtg_switch_txt_crr, 
levels=c("0", "1", "2", "3", "4"),
labels = c("No DTG", "DTG switch", "Death", "LTFU", "Transfer"))

FU_time$dtg_switch_txt_crr1b.factor <- factor(FU_time$dtg_switch_txt_crr1b, 
levels=c("0", "1", "2", "3", "4"),
labels = c("a", "b", "c", "d", "e"))

FU_time$dtg_switch_txt_crr2.factor <- factor(FU_time$dtg_switch_txt_crr2,
levels=c("0", "1", "2"),
labels = c("a", "c", "b"))

FU_time$dtg_switch_txt_crr2b.factor <- factor(FU_time$dtg_switch_txt_crr2,
levels=c("0", "1", "2"),
labels = c("RIC", "DTG switch", "competing risks"))

FU_time$dtg_switch_txt_crr3b.factor <- factor(FU_time$dtg_switch_txt_crr3,
levels=c("0", "1"),
labels = c("No Switch", "DTG switch"))



frq(FU_time, dtg_switch_txt_os)
frq(FU_time, dtg_switch_txt_dss)
frq(FU_time, dtg_switch_txt_crr)
frq(FU_time, dtg_switch_txt_crr.factor)
frq(FU_time, dtg_switch_txt_crr2)
frq(FU_time, dtg_switch_txt_crr2.factor)
frq(FU_time, dtg_switch_txt_crr2b.factor)
frq(FU_time, dtg_switch_txt_crr3)
frq(FU_time, dtg_switch_txt_crr3b.factor)
   
```

```{r}
# Label and recode other variables
FU_time <- FU_time %>%
 mutate(sex2 = factor(sex) %>% 
 fct_recode("1" = "Male",
            "0" = "Female")) %>%
  mutate(dtg.factor = factor(dtg_switch))

```



<!-- ----------------------------------------------------- -->


# National DTG adoption dates

Pulled by @aezaniewski from different sources.  

```{r include=FALSE}
dtg_recomend = data.table(read_fst("data_temp/dtg_recomend_v1a.fst"))

dtg_recomend$ctry <- factor(dtg_recomend$country, levels = c( "LSO","MWI", "MOZ" ,"ZAF", "ZMB", "ZWE"),
 labels = c("Lesotho",  "Malawi", "Mozambique", "South Africa", "Zambia", "Zimbabwe"))



```

```{r echo=FALSE}
dtg_recomend %>% select(country, ctry, dtg_recomend_d_initiate, dtg_recomend_d) %>%
  kbl() %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  column_spec(1, italic = TRUE) %>%
  row_spec(1, italic = FALSE)
```

<!-- ----------------------------------------------------- -->


<!-- ----------------------------------------------------- -->

# Transition cohort: baseline characteristics


## Country by sex

```{r}
sjPlot::tab_xtab(tblBAS_2$ctry,
                 tblBAS_2$sex,
                 show.row.prc = TRUE, show.summary = FALSE)
```

## Country by sex & age

```{r}
sjPlot::tab_xtab(tblBAS_2$ctry,
                 tblBAS_2$sex_age2_txt,
                 show.row.prc = TRUE, show.summary = FALSE)
```



<!-- ----------------------------------------------------- -->

# Median time on ART (months)


```{r}

tblBAS_2b <- tblBAS_2 %>% mutate(timeonART_d = as.numeric(dtg_recomend_d - recart_d)) %>%
  mutate(timeonART_m = timeonART_d/30.5) 
  

# overall
timeonART_sex <- tblBAS_2b %>% 
  group_by(sex_age2_txt) %>% 
   filter(!is.na(timeonART_m)) %>%  
  summarise(#"quantile10"= round(quantile(timeonART_m,c(.10)),digits=1),
            "Median"= round(quantile(timeonART_m,c(.50)),digits=1),
            "quantile25"= round(quantile(timeonART_m,c(.25)),digits=1),
            "quantile75"= round(quantile(timeonART_m,c(.75)),digits=1),
            #"quantile90"= round(quantile(timeonART_m,c(.90)),digits=1),
            #"Mean" = round(mean(timeonART_m),digits=1)
            )
  

# by country
timeonART_ctry_sex <- tblBAS_2b %>% 
  group_by(ctry, country, sex_age2_txt) %>% 
   filter(!is.na(timeonART_m)) %>%  
   summarise(#"quantile10"= round(quantile(timeonART_m,c(.10)),digits=1),
            "Median"= round(quantile(timeonART_m,c(.50)),digits=1),
            "quantile25"= round(quantile(timeonART_m,c(.25)),digits=1),
            "quantile75"= round(quantile(timeonART_m,c(.75)),digits=1),
            #"quantile90"= round(quantile(timeonART_m,c(.90)),digits=1),
            #"Mean" = round(mean(timeonART_m),digits=1)
            )

```

## Overall

```{r}

writeLines("td, th { padding : 6px } th { background-color : brown ; color : white; border : 1px solid white; } td { color : brown ; border : 1px solid brown }", con = "mystyle.css")
dset1 <- head(timeonART_sex)
dset1 <- (timeonART_sex)
knitr::kable(dset1, format = "html")

```

```{r eval=FALSE, include=FALSE}
table1::table1(~ Median + quantile25 + quantile75 | sex_age2_txt , data = timeonART_sex, overall=c(left="Total"), render.categorical = my.render.cat)
```

## Lesotho

```{r echo=FALSE}

timeonART_ctry_sex2 <- timeonART_ctry_sex %>% filter(country =="LSO") 

writeLines("td, th { padding : 6px } th { background-color : brown ; color : white; border : 1px solid white; } td { color : brown ; border : 1px solid brown }", con = "mystyle.css")
dset1 <- head(timeonART_ctry_sex2)
dset1 <- (timeonART_ctry_sex2)
knitr::kable(dset1, format = "html")

```


## Malawi

```{r echo=FALSE}

timeonART_ctry_sex2 <- timeonART_ctry_sex %>% filter(country =="MWI") 

writeLines("td, th { padding : 6px } th { background-color : brown ; color : white; border : 1px solid white; } td { color : brown ; border : 1px solid brown }", con = "mystyle.css")
dset1 <- head(timeonART_ctry_sex2)
dset1 <- (timeonART_ctry_sex2)
knitr::kable(dset1, format = "html")

```



## Mozambique

```{r echo=FALSE}

timeonART_ctry_sex2 <- timeonART_ctry_sex %>% filter(country =="MOZ") 

writeLines("td, th { padding : 6px } th { background-color : brown ; color : white; border : 1px solid white; } td { color : brown ; border : 1px solid brown }", con = "mystyle.css")
dset1 <- head(timeonART_ctry_sex2)
dset1 <- (timeonART_ctry_sex2)
knitr::kable(dset1, format = "html")

```


## South Africa

```{r echo=FALSE}

timeonART_ctry_sex2 <- timeonART_ctry_sex %>% filter(country =="ZAF") 

writeLines("td, th { padding : 6px } th { background-color : brown ; color : white; border : 1px solid white; } td { color : brown ; border : 1px solid brown }", con = "mystyle.css")
dset1 <- head(timeonART_ctry_sex2)
dset1 <- (timeonART_ctry_sex2)
knitr::kable(dset1, format = "html")

```


## Zambia

```{r echo=FALSE}

timeonART_ctry_sex2 <- timeonART_ctry_sex %>% filter(country =="ZMB") 

writeLines("td, th { padding : 6px } th { background-color : brown ; color : white; border : 1px solid white; } td { color : brown ; border : 1px solid brown }", con = "mystyle.css")
dset1 <- head(timeonART_ctry_sex2)
dset1 <- (timeonART_ctry_sex2)
knitr::kable(dset1, format = "html")

```



## Zimbabwe

```{r echo=FALSE}

timeonART_ctry_sex2 <- timeonART_ctry_sex %>% filter(country =="ZWE") 

writeLines("td, th { padding : 6px } th { background-color : brown ; color : white; border : 1px solid white; } td { color : brown ; border : 1px solid brown }", con = "mystyle.css")
dset1 <- head(timeonART_ctry_sex2)
dset1 <- (timeonART_ctry_sex2)
knitr::kable(dset1, format = "html")

```





<!-- ----------------------------------------------------- -->

# Transition cohort: Switching to DTG

Number of adults on ART switching to "DTG ART".

## by country

```{r}
sjPlot::tab_xtab(tblBAS_2$ctry,
                 tblBAS_2$dtg_switch_txt,
                 show.row.prc = TRUE, show.summary = FALSE)
```


## by sex & age

```{r}
sjPlot::tab_xtab(tblBAS_2$sex_age2_txt,
                 tblBAS_2$dtg_switch_txt,
                 show.row.prc = TRUE, show.summary = FALSE)
```


<!-- ----------------------------------------------------- -->

# Transition cohort: figures


```{r include=FALSE}
frq(FU_time, dtg_recomend_d)
frq(FU_time, dtg_recomend_t)
```


## Fig 3: Cumulative incidence (1-KM plots)

By country and sex by calendar year, **NOT taking into account competing risks**.


```{r}
#ggsurvplot

FU_time_temp <- FU_time 

FU_time_temp$sex_age2_txt <- factor(FU_time_temp$sex_age2_txt, levels = c("Male 20-49", "Male 50+", "Female 20-49", "Female 50+"),
                       labels = c( "Men 20-49", "Men 50+", "Women 20-49", "Women 50+"))


fill <- c("darkblue", "#56B4E9", "#ac0200", "#fcc8c7" )

fit_t0 = survfit2(Surv(time_t2, dtg_switch) ~ sex_age2_txt , data = FU_time_temp)

#put model results into a dataframe
results <- broom::tidy(fit_t0)

#fit_t0 = survfit2(Surv(time_t2, dtg_switch_txt_crr3) ~ sex , data = FU_time)


myplot <- ggsurvplot(fit_t0, data = FU_time_temp, fun = "event", size = 1.5,
           censor = FALSE,
           risk.table = FALSE,
           facet.by = "ctry",
           # to remove the heading within the facet.wrap panels
           short.panel.labs = TRUE, 
           ncol = 2, 
           surv.scale = "percent",
           xlab = "", 
           ylab = "Cumulative incidence (%)",
           title = "Switching to dolutegravir", 
           xlim = c(0, 1643), break.x.by = 365,
           legend.title="")

#myplot 

plot2 <- myplot +
  theme_ggsurvfit_default() +
      #scale_x_continuous(breaks = c(153, 518, 883, 1248, 1613 ), limits = c(0,1613), 
    scale_x_continuous(breaks = c(183, 548, 913, 1278, 1643 ), limits = c(0,1643),
    labels = c("2019", "2020", "2021", "2022", "2023")) +
    scale_color_manual(values = c("darkblue", "#56B4E9", "#ac0200", "#fcc8c7")) +
  theme(
    axis.title = element_text(size = 12),
          plot.title = element_text(size = 14, hjust = 0.5), #center title
          axis.title.y = element_text(margin = margin(t = 1, r = 1, b = 1, l = 1), size = 12), #pad y axis label
          axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0), size = 12), #pad x axis label
          axis.text.x = element_text(size = 12, angle =90, vjust = 0.6, hjust = 0.5),
          axis.text.y = element_text(size = 10),
          axis.text = element_text(size = 12),
          legend.text = element_text(size = 12),
          legend.title = element_text(size = 12),
          legend.position = "right", 
          legend.justification = "center",
          legend.box.background = element_rect(color = "black"),
          legend.direction="vertical", 
          legend.margin=margin(0,0,0,0),
          legend.box.margin = margin(-5, -5, -5, -5), 
          legend.key.size = unit(0.7, "cm"), 
          strip.background = element_blank(),
         strip.text.x = element_text(size = 12)
  ) +
  guides(color = guide_legend(ncol = 1, reverse = FALSE))  +
  geom_vline(data = FU_time, aes(xintercept = dtg_recomend_t),linetype="dashed") 
plot2


ggsave("output/Paper/Fig3_switch_KM_ctry_calendar.png", width = 7, height =8)

```


## Plot with at risk table

```{r}
sf <- survfit2(Surv(time_t2, dtg_switch) ~ ctry, data = FU_time_temp)

p <- sf %>% 
  ggsurvfit(type = "risk", linetype_aes = TRUE, linewidth = 1) +
  labs(
    x = "",
    y = "Cumulative incidence"
  ) +
  add_confidence_interval() +
  annotate("text", x = 1000, y = 0.05, label = glue::glue("{survfit2_p(sf)}")) +
  add_risktable()

  
  # add_censor_mark() +
  # add_confidence_interval() +
  # add_quantile() +
 #  add_risktable()
 # add_pvalue()


## edit with ggplot

plot3 <- p + 
  theme_ggsurvfit_default() +
  # limit plot to show 8 years and less
  coord_cartesian(xlim = c(0, 1643), clip = 'off', expand=TRUE) +
    theme(legend.position = "bottom") +
  # update figure labels/titles
  labs(
    y = "Cumulative incidence",
    title = "1-KM switching to DTG",
  ) +
  # reduce padding on edges of figure and format axes
  scale_y_continuous(label = scales::percent, 
                     breaks = seq(0, 1, by = 0.2),
                     expand = c(0.015, 0)) +
       #scale_x_continuous(breaks = c(153, 518, 883, 1248, 1613 ), limits = c(0,1613), 
    scale_x_continuous(breaks = c(0, 183, 548, 913, 1278, 1643 ), limits = c(0,1643),
    labels = c("start", "2019", "2020", "2021", "2022", "2023")) +
  guides(colour=guide_legend(ncol=6))

plot3

ggsave("output/Paper/Fig3_switch_at_risk_table_ctry_calendar.png", width = 8, height =8)

```

```{r include=FALSE}
with(sf, data.frame(time, n.risk, n.event, surv, cumhaz))


```


<!-- ----------------------------------------------------- -->

# Output of KM model 

Print out "survival" at end of plot time_t2 ==1643 (end 2022)


```{r include=FALSE}

sf2 <- survfit2(Surv(time_t2, dtg_switch) ~ ctry + sex_age2_txt , data = FU_time_temp)

library(broom)

members <- c("time", "n.risk", "n.censor", "n.event", "strata") 

#create tidy dataframe and subset by the columns saved in members
df_sf2 <- tidy(sf2)[,members]
head(df_sf2)

df_sf2_2023 <- df_sf2 %>% filter(time >1610 & time <1645)
```


```{r eval=FALSE, include=FALSE}


writeLines("td, th { padding : 6px } th { background-color : brown ; color : white; border : 1px solid white; } td { color : brown ; border : 1px solid brown }", con = "mystyle.css")
dset1 <- head(df_sf2_2023)
knitr::kable(dset1, format = "html")
```


summary(survfit2(Surv(time_t2, dtg_switch) ~ ctry + sex_age2_txt, data = FU_time_temp), times = 1643)

## Lesotho

```{r echo=FALSE}
#Print out "survival" at end of plot time_t2 ==1643 (end 2022)
FU_time_temp2 <- FU_time_temp %>% filter(country =="LSO")

summary(survfit2(Surv(time_t2, dtg_switch) ~ ctry + sex_age2_txt  , data = FU_time_temp2), times = 1643)
```


## Malawi

```{r echo=FALSE}
#Print out "survival" at end of plot time_t2 ==1643 (end 2022)
FU_time_temp2 <- FU_time_temp %>% filter(country =="MWI")

summary(survfit2(Surv(time_t2, dtg_switch) ~ ctry + sex_age2_txt  , data = FU_time_temp2), times = 1643)
```


## Mozambique

```{r echo=FALSE}
#Print out "survival" at end of plot time_t2 ==1643 (end 2022)
FU_time_temp2 <- FU_time_temp %>% filter(country =="MOZ")

summary(survfit2(Surv(time_t2, dtg_switch) ~ ctry + sex_age2_txt  , data = FU_time_temp2), times = 1643)
```


## South Africa

```{r echo=FALSE}
#Print out "survival" at end of plot time_t2 ==1643 (end 2022)
FU_time_temp2 <- FU_time_temp %>% filter(country =="ZAF")

summary(survfit2(Surv(time_t2, dtg_switch) ~ ctry + sex_age2_txt  , data = FU_time_temp2), times = 1643)
```


## Zambia

```{r echo=FALSE}
#Print out "survival" at end of plot time_t2 ==1643 (end 2022)
FU_time_temp2 <- FU_time_temp %>% filter(country =="ZMB")

summary(survfit2(Surv(time_t2, dtg_switch) ~ ctry + sex_age2_txt  , data = FU_time_temp2), times = 1643)
```


## Zimbabwe

```{r echo=FALSE}
#Print out "survival" at end of plot time_t2 ==1643 (end 2022)
FU_time_temp2 <- FU_time_temp %>% filter(country =="ZWE")

summary(survfit2(Surv(time_t2, dtg_switch) ~ ctry + sex_age2_txt  , data = FU_time_temp2), times = 1643)
```



<!-- ----------------------------------------------------- -->

# Computing Environment

```{r echo=FALSE, results='asis'}
report(sessionInfo())
```