---
title: "IeDEA SA256 DTG uptake: 02 Abstract Outcomes Fig 1"
subtitle: " "
author: "Elizabeth Zaniewski,Radek Panczak, Eliane Rohner"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    code_folding: show
    self_contained: true
    highlight: pygments
    keep_md: no
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
---

```{r r-setup, include = FALSE}
options(scipen = 999)
options(max.print = "300")
set.seed(12345)

library(pacman)
p_load(sf, tidyverse, kableExtra, report, magrittr, fst, data.table, survival, lubridate, ggsurvfit, gtsummary, tidycmprsk, Hmisc, survminer, survMisc)

p_load(
  kableExtra,
  scales, ggplot2, dplyr,
  fst, data.table, sjmisc,
  osfr
)

#install.packages("finalfit")

library(survival)
library(lubridate)
library(ggsurvfit)
library(gtsummary)
library(tidycmprsk)
library(Hmisc)
library(survminer)
library(forcats)
library(dplyr)
#library(condsurv)
library(patchwork)



import::from("sjmisc", "frq")
import::from("gmodels", "CrossTable")
```

```{r conflicts, include = FALSE}
tidyverse::tidyverse_conflicts()

conflicted::conflict_scout()
```

```{r knit-setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

knitr::opts_chunk$set(cache = FALSE,
                      prompt = FALSE,
                      tidy = FALSE,
                      comment = NA,
                      message = FALSE,
                      warning = FALSE,
                      echo = FALSE)

knitr::opts_knit$set(width = 200)
```

<!-- ----------------------------------------------------- -->

```{r}
#Clear existing data and graphics
rm(list=ls())
graphics.off()
```

```{r}

#####################################################################
# to adjust decimal places of categorical variables

my.render.cat <- function(x) {
  c("", sapply(stats.default(x), function(y) with(y,
                                                  sprintf("%d (%0.0f %%)", FREQ, PCT))))
}

#####################################################################

```



<!-- ----------------------------------------------------- -->


# National DTG adoption dates

Pulled by @aezaniewski from different sources.  

```{r include=FALSE}
dtg_recomend = data.table(read_fst("data_temp/dtg_recomend_v1a.fst"))

dtg_recomend$ctry <- factor(dtg_recomend$country, levels = c( "LSO","MWI", "MOZ" ,"ZAF", "ZMB", "ZWE"),
 labels = c("Lesotho",  "Malawi", "Mozambique", "South Africa", "Zambia", "Zimbabwe"))



```

```{r echo=FALSE}
dtg_recomend %>% select(country, ctry, dtg_recomend_d_initiate, dtg_recomend_d) %>%
  kbl() %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  column_spec(1, italic = TRUE) %>%
  row_spec(1, italic = FALSE)
```

<!-- ----------------------------------------------------- -->

```{r include=FALSE}
tblBAS_1 = data.table(read_fst("data_temp/tblBAS_initiation_v2_outcome.fst"))

tblBAS_1$ctry <- factor(tblBAS_1$country, levels = c( "LSO","MWI", "MOZ" ,"ZAF", "ZMB", "ZWE"),
 labels = c("Lesotho",  "Malawi", "Mozambique", "South Africa", "Zambia", "Zimbabwe"))


```


```{r include=FALSE}
tblBAS_2 = data.table(read_fst("data_temp/tblBAS_transition_v2_outcome.fst"))

```

```{r}
tblBAS_1 <- tblBAS_1 %>%
  mutate(sex_age2_txt = ifelse(sex_age2 =="Male.[20,50)", "Male 20-49",
                               ifelse(sex_age2 =="Female.[20,50)", "Female 20-49",
                                    ifelse(sex_age2 =="Male.[50,100]", "Male 50+",
                                         ifelse(sex_age2 =="Female.[50,100]", "Female 50+", "other"
                                      )))))

tblBAS_1 <- tblBAS_1 %>%
  mutate(sex_age3_txt = ifelse(sex_age3 =="Male.[20,30)", "Male 20-29",
                               ifelse(sex_age3 =="Female.[20,30)", "Female 20-29",
                                ifelse(sex_age3 =="Male.[30,40)", "Male 30-39",
                               ifelse(sex_age3 =="Female.[30,40)", "Female 30-39",
                              ifelse(sex_age3 =="Male.[40,50)", "Male 40-49",
                               ifelse(sex_age3 =="Female.[40,50)", "Female 40-49",
                                ifelse(sex_age3 =="Male.[50,100]", "Male 50+",
                                ifelse(sex_age3 =="Female.[50,100]", "Female 50+", "other"
                                      )))))))))

tblBAS_2 <- tblBAS_2 %>%
  mutate(sex_age2_txt = ifelse(sex_age2 =="Male.[20,50)", "Male 20-49",
                               ifelse(sex_age2 =="Female.[20,50)", "Female 20-49",
                                    ifelse(sex_age2 =="Male.[50,100]", "Male 50+",
                                         ifelse(sex_age2 =="Female.[50,100]", "Female 50+", "other"
                                      )))))


tblBAS_2 <- tblBAS_2 %>%
  mutate(sex_age3_txt = ifelse(sex_age3 =="Male.[20,30)", "Male 20-29",
                               ifelse(sex_age3 =="Female.[20,30)", "Female 20-29",
                                ifelse(sex_age3 =="Male.[30,40)", "Male 30-39",
                               ifelse(sex_age3 =="Female.[30,40)", "Female 30-39",
                              ifelse(sex_age3 =="Male.[40,50)", "Male 40-49",
                               ifelse(sex_age3 =="Female.[40,50)", "Female 40-49",
                                ifelse(sex_age3 =="Male.[50,100]", "Male 50+",
                                ifelse(sex_age3 =="Female.[50,100]", "Female 50+", "other"
                                      )))))))))
```

<!-- ----------------------------------------------------- -->

# Overview

**For adults aged 20-39 years only!!**

```{r}
#only keep adults aged 20-39 years
tblBAS_1 <- tblBAS_1 %>% filter(age_cat3 == "[20,30)" | age_cat3 =="[30,40)")
tblBAS_2 <- tblBAS_2 %>% filter(age_cat3 == "[20,30)" | age_cat3 =="[30,40)")
```


## Initiation cohort

```{r}

frq(tblBAS_1, sex)

temp1 <- tblBAS_1 %>% select(country, program, dtg_recomend_d, max_close_d) %>%
  arrange(dtg_recomend_d, country, program) %>%
  group_by(country, program, dtg_recomend_d, max_close_d) %>%
  mutate(Npatients = n()) %>%
  distinct()
  
temp1 %>% 
  kbl() %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  column_spec(1, italic = TRUE) %>%
  row_spec(1, italic = FALSE)

```

## Transition cohort

```{r}
frq(tblBAS_2, sex)


temp2 <- tblBAS_2 %>% select(country, program, dtg_recomend_d, max_close_d) %>%
  arrange(dtg_recomend_d, country, program) %>%
  group_by(country, program, dtg_recomend_d, max_close_d) %>%
  mutate(Npatients = n()) %>%
  distinct()
  
temp2 %>% 
  kbl() %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  column_spec(1, italic = TRUE) %>%
  row_spec(1, italic = FALSE)
```



```{r echo=FALSE}
table1::label(tblBAS_1$sex) <- "Sex"
table1::label(tblBAS_1$age_cat3) <- "Age group"
table1::label(tblBAS_1$dtg_first_txt) <- "ART at initiation"
table1::label(tblBAS_1$sex_age2) <- "Sex & age group"
table1::label(tblBAS_1$sex_age3) <- "Sex & age group"
table1::label(tblBAS_1$program) <- "Program"
table1::label(tblBAS_1$ctry) <- "Country"
table1::label(tblBAS_1$sex_age2_txt) <- "Sex & age group"
table1::label(tblBAS_1$sex_age3_txt) <- "Sex & age group"

```


```{r}
table1::label(tblBAS_2$sex) <- "Sex"
table1::label(tblBAS_2$age_cat3) <- "Age group"
table1::label(tblBAS_2$dtg_switch_txt) <- "DTG switch"
table1::label(tblBAS_2$sex_age2) <- "Sex & age group"
table1::label(tblBAS_2$sex_age3) <- "Sex & age group"
table1::label(tblBAS_2$program) <- "Program"
table1::label(tblBAS_2$ctry) <- "Country"
table1::label(tblBAS_2$sex_age2_txt) <- "Sex & age group"
table1::label(tblBAS_2$sex_age3_txt) <- "Sex & age group"
```


<!-- ----------------------------------------------------- -->

# Initiation Cohort: Outcomes

```{r eval=FALSE, include=FALSE}

frq(tblBAS_1, dtg_first)

tblBAS_1[, start_fu := recart_d] 
tblBAS_1[, end_fu := rev_outcome_d] 

tblBAS_1[, FU_day := as.numeric(end_fu - start_fu)]
tblBAS_1[, FU_month := round(as.numeric(end_fu - start_fu) / (365.25 / 12), 1)]
tblBAS_1[, FU_year := round(as.numeric(end_fu - start_fu) / (365.25), 1)]

tblBAS_1[, time := FU_day]

```


```{r}
table1::label(tblBAS_1$sex) <- "Sex"
table1::label(tblBAS_1$age_cat3) <- "Age group"

table1::label(tblBAS_1$rev_outcome) <- "Outcome"
table1::label(tblBAS_1$sex_age2) <- "Sex & age group"
table1::label(tblBAS_1$sex_age3) <- "Sex & age group"
table1::label(tblBAS_1$program) <- "Program"
table1::label(tblBAS_1$ctry) <- "Country"

```


```{r}
FU_time_1 <- tblBAS_1 %>% 
  select(patient, program, country, ctry, dtg_recomend_d, dtg, dtg_final, dtg_first, dtg_first_txt, max_close_d, rev_outcome, rev_outcome_d, start_fu, end_fu, time) %>%
  mutate(start_min_d = min(start_fu)) %>% # finds min start date across all cohorts (=dtg adoption date)
  mutate(time_t2 = as.numeric(end_fu) - as.numeric(start_min_d)) %>% # gives end time relative to earliest start min date (=earliest dtg adoption date)
  mutate(dtg_recomend_t = as.numeric(dtg_recomend_d) - as.numeric(start_min_d)) %>% # gives dtg adoption date relative to earliest dtg adoption date
  mutate(dtg_comp_b = ifelse(rev_outcome == "RIC", 0, 1))

# create file of dtg_recomend_t
dtg_recomend_t <- FU_time_1 %>% select(program, country, ctry, dtg_recomend_d, dtg_recomend_t) %>% unique()

#dtg_LSO <- dtg_recomend_t %>% filter(country =="LSO")
#dtg_MWI <- dtg_recomend_t %>% filter(country =="MWI")
#dtg_MOZ <- dtg_recomend_t %>% filter(country =="MOZ")
#dtg_ZAF <- dtg_recomend_t %>% filter(country =="ZAF")
#dtg_ZMB <- dtg_recomend_t %>% filter(country =="ZMB")
#dtg_ZWE <- dtg_recomend_t %>% filter(country =="ZWE")


```

```{r include=FALSE}
 # Competing risks regression (alive ==0, dtg ==1, death ==2, ltfu ==3, trans ==4)
  
frq(FU_time_1, rev_outcome)


FU_time_1 <- FU_time_1 %>% 
  mutate(
    
dtg_first_txt_crr1b = case_when(
    rev_outcome == "RIC" ~ 0, # still alive RIC (retained in care)
    rev_outcome == "Death" ~ 1, # outcome Death
    rev_outcome == "Transfer" ~ 2, # outcome Transfer
    rev_outcome == "LTFU" ~ 3), # outcome LTFU
)

# factor alphabetically, so can determine plot output order of events (_crr1b.factor)
FU_time_1$dtg_first_txt_crr1b.factor <- factor(FU_time_1$dtg_first_txt_crr1b, 
levels=c("0", "1", "2", "3"),
labels = c("a", "b", "c", "d"))

```


```{r include=FALSE}
# create two subcohorts: intitated DTG & did not initiate DTG
frq(FU_time_1, dtg_final)
frq(FU_time_1, rev_outcome)

FU_time_1_DTG <- FU_time_1 %>% filter(dtg_first == 1)
FU_time_1_noDTG <- FU_time_1 %>% filter(dtg_first == 0)

```


## All adults 20-39 years

Initiated DTG ART

```{r echo=FALSE}

## by country with DTG
sjPlot::tab_xtab(FU_time_1_DTG$ctry,
                 FU_time_1_DTG$rev_outcome,
                 show.row.prc = TRUE, show.summary = FALSE)

```

Initiated other ART


```{r echo=FALSE}

## by country with no DTG
sjPlot::tab_xtab(FU_time_1_noDTG$ctry,
                 FU_time_1_noDTG$rev_outcome,
                 show.row.prc = TRUE, show.summary = FALSE)
```

<!-- ----------------------------------------------------- -->

# Plots for initiation cohort

## Initiated DTG ART by country

Cumulative incidence plot of outcomes among adults initiating DTG ART.

```{r}

## Need to fix issue with LSO, ZAF, MOZ not having anyone intiating DTG early...
testLSO <- FU_time_1_DTG %>% filter(country =="LSO") %>% filter(rev_outcome == "RIC")
testZAF <- FU_time_1_DTG %>% filter(country =="ZAF") %>% filter(rev_outcome == "RIC")
testMOZ <- FU_time_1_DTG %>% filter(country =="MOZ") %>% filter(rev_outcome == "RIC")
testMWI <- FU_time_1_DTG %>% filter(country =="MWI") %>% filter(rev_outcome == "RIC")
testZMB <- FU_time_1_DTG %>% filter(country =="ZMB") %>% filter(rev_outcome == "RIC")
testZWE <- FU_time_1_DTG %>% filter(country =="ZWE") %>% filter(rev_outcome == "RIC")

```

```{r}
FU_time_1_DTG <- FU_time_1_DTG %>% mutate(time_t2 = ifelse(patient =="SMLM2152", 124, time_t2))
FU_time_1_DTG <- FU_time_1_DTG %>% mutate(time_t2 = ifelse(patient =="HLHL000134", 458, time_t2))
FU_time_1_DTG <- FU_time_1_DTG %>% mutate(time_t2 = ifelse(patient =="SMMA5231", 305, time_t2))
FU_time_1_DTG <- FU_time_1_DTG %>% mutate(time_t2 = ifelse(patient =="LH10037287", 185, time_t2))
FU_time_1_DTG <- FU_time_1_DTG %>% mutate(time_t2 = ifelse(patient =="I_SA_CIDRZ_000203-05", 1, time_t2))
FU_time_1_DTG <- FU_time_1_DTG %>% mutate(time_t2 = ifelse(patient =="SMZC13476", 305, time_t2))
```


```{r}
 rm(testLSO, testMOZ, testMWI, testZAF, testZMB, testZWE)
```


```{r include=FALSE}

fit1_dtg1_ctry <- survfit2(Surv(time_t2, dtg_comp_b) ~ ctry, data = FU_time_1_DTG, etype = dtg_first_txt_crr1b.factor)

#ggcompetingrisks(fit = fit1_ctry_dtg) + 
 #facet_wrap(~strata, ncol = 1)

ggcompetingrisks(fit = fit1_dtg1_ctry) + 
 facet_wrap(~strata, ncol = 1) +
  geom_vline(data = FU_time_1_DTG, aes(xintercept = dtg_recomend_t),linetype="dashed", size = 0.5) 
  

  #geom_vline(xintercept = c(123, 184) ,linetype="dashed", size = 0.5) +
  #annotate("text", x=123, y=0.8, label="Leotho", angle=90, size = 4, hjust = 0.9, vjust = 0.9) +
  #annotate("text", x=184, y=0.8, label="Malawi", angle=90, size = 4, hjust = 0.9, vjust = 0.9)

#ggcompetingrisks(fit = fit1_ctry_dtg) + 
 #facet_wrap(~strata, ncol = 3)
```


```{r}

p05 <-ggcompetingrisks(fit = fit1_dtg1_ctry,
                 xlab = "", 
                  ylab = "Cumulative incidence (%)",
                 title = "") +
              #scale_x_continuous(breaks = c(153, 518, 883, 1248, 1613 ), limits = c(0,1613), 
    scale_x_continuous(breaks = c(183, 548, 913, 1278, 1643 ), limits = c(0,1643), 
    labels = c("2019", "2020", "2021", "2022", "2023")) +
  theme_ggsurvfit_default() +
 scale_fill_manual("",
  values = c("#9ef0f0", "gray", "#08bdba",  "purple","#0f62fe"), 
  labels = c("RIC", "Death", "Transfer","LTFU"),  
  name = "Outcome event") +
 facet_wrap(~strata, ncol = 1, labeller = labeller(strata = 
      c("ctry=Lesotho" = "Lesotho",
         "ctry=Malawi" = "Malawi", 
        "ctry=Mozambique" = "Mozambique", 
        "ctry=South Africa" = "South Africa",
        "ctry=Zambia" = "Zambia", 
         "ctry=Zimbabwe" = "Zimbabwe"
))) 

#p05

plot1_dtg1_ctry <- p05 + theme(
          plot.title = element_text(size = 10, hjust = 0.5), #center title
          plot.subtitle = element_text(size = 10, hjust = 0.5), #center subtitle
          axis.title.y = element_text(margin = margin(t = 1, r = 1, b = 1, l = 1), size = 11), #pad y axis label
          axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0), size = 11), #pad x axis label
          axis.text.x = element_text(size = 11, angle=90,vjust = 0.6, hjust = 0.5),
          axis.text.y = element_text(size = 11),
          legend.text = element_text(size = 11),
          legend.title = element_text(size = 11),
          legend.position = "right", 
          legend.justification = "left",
          legend.box.background = element_rect(color = "black"),
          legend.direction="vertical", 
          legend.margin=margin(0,0,0,0),
          legend.box.margin = margin(-5, -5, -5, -5), 
          legend.key.size = unit(0.7, "cm"),
          #strip.background = element_blank(),
          strip.text.x = element_text(size = 12)
          ) +
  guides(color = guide_legend(ncol = 2))  +
    scale_y_continuous(breaks = c(0, 0.25,  0.50, 0.75, 1.0), 
                        labels = c("0%", "25%", "50%", "75%", "100%")) 

plot1_dtg1_ctry

ggsave("output/Abstract_outcomes/Fig1_cohort1_dtg1_outcomes_ctry_calendar.png", width = 5, height =12)
```

```{r eval=FALSE, include=FALSE}
+
  geom_vline(data = dtg_recomend_t, aes(xintercept = dtg_recomend_t),linetype="dashed")  +
  annotate("text", x=123, y=0.8, label="Leotho", angle=90, size = 4, hjust = 0.9, vjust = 0.9) +
  annotate("text", x=184, y=0.8, label="Malawi", angle=90, size = 4, hjust = 0.9, vjust = 0.9)
```


## Initiated non-DTG ART by country


```{r}

## Need to fix issue with LSO, ZAF, MOZ not having anyone intiating DTG early...
testLSO <- FU_time_1_noDTG %>% filter(country =="LSO") %>% filter(rev_outcome == "RIC")
testMWI <- FU_time_1_noDTG %>% filter(country =="MWI") %>% filter(rev_outcome == "RIC")
testMOZ <- FU_time_1_noDTG %>% filter(country =="MOZ") %>% filter(rev_outcome == "RIC")
testZAF <- FU_time_1_noDTG %>% filter(country =="ZAF") %>% filter(rev_outcome == "RIC")
testZMB <- FU_time_1_noDTG %>% filter(country =="ZMB") %>% filter(rev_outcome == "RIC")
testZWE <- FU_time_1_noDTG %>% filter(country =="ZWE") %>% filter(rev_outcome == "RIC")

```

```{r}
test2 <- FU_time_1_noDTG %>% filter(country =="ZWE") 
```


```{r}
FU_time_1_noDTG <- FU_time_1_noDTG %>% mutate(time_t2 = ifelse(patient =="SMLM2266", 124, time_t2))
FU_time_1_noDTG <- FU_time_1_noDTG %>% mutate(time_t2 = ifelse(patient =="LH10072522", 185, time_t2))
FU_time_1_noDTG <- FU_time_1_noDTG %>% mutate(time_t2 = ifelse(patient =="SMMA4956", 305, time_t2))
FU_time_1_noDTG <- FU_time_1_noDTG %>% mutate(time_t2 = ifelse(patient =="HLHL000801", 458, time_t2))
FU_time_1_noDTG <- FU_time_1_noDTG %>% mutate(time_t2 = ifelse(patient =="I_SA_CIDRZ_000038-05", 1, time_t2))
FU_time_1_noDTG <- FU_time_1_noDTG %>% mutate(time_t2 = ifelse(patient =="SMZC13247", 305, time_t2))
```

```{r}
 rm(testLSO, testMOZ, testMWI, testZAF, testZMB, testZWE)
```


Cumulative incidence plot of outcomes among adults initiating non-DTG ART.


```{r include=FALSE}

fit1_dtg0_ctry <- survfit2(Surv(time_t2, dtg_comp_b) ~ ctry, data = FU_time_1_noDTG, etype = dtg_first_txt_crr1b.factor)

fit1_dtg0_ctry_v2 <- survfit2(Surv(time_t2, dtg_first_txt_crr1b.factor) ~ ctry, data = FU_time_1_noDTG)

ggcompetingrisks(fit = fit1_dtg0_ctry_v2) + 
 facet_wrap(~strata, ncol = 1) +
  geom_vline(data = FU_time_1_DTG, aes(xintercept = dtg_recomend_t),linetype="dashed", size = 0.5) +
      scale_x_continuous(breaks = c(0, 500, 1000,1500, 2000), limits = c(0,2000), 
    labels = c("0", "500", "1000", "1500", "2000")) 


#ggcompetingrisks(fit = fit1_ctry_dtg) + 
 #facet_wrap(~strata, ncol = 1)

ggcompetingrisks(fit = fit1_dtg0_ctry) + 
 facet_wrap(~strata, ncol = 1) +
  geom_vline(data = FU_time_1_DTG, aes(xintercept = dtg_recomend_t),linetype="dashed", size = 0.5) +
      scale_x_continuous(breaks = c(0, 500, 1000,1500, 2000), limits = c(0,2000), 
    labels = c("0", "500", "1000", "1500", "2000")) 

#ggcompetingrisks(fit = fit1_ctry_dtg) + 
 #facet_wrap(~strata, ncol = 3)
```


```{r}

p05 <-ggcompetingrisks(fit = fit1_dtg0_ctry_v2,
                 xlab = "Years since ART initiation", 
                  ylab = "Cumulative incidence (%)",
                 title = "") +
              #scale_x_continuous(breaks = c(153, 518, 883, 1248, 1613 ), limits = c(0,1613), 
    scale_x_continuous(breaks = c(183, 548, 913, 1278, 1643 ), limits = c(0,1643),  
    labels = c("2019", "2020", "2021", "2022", "2023")) +
  theme_ggsurvfit_default() +
 scale_fill_manual("",
  values = c("#9ef0f0", "gray", "#08bdba",  "purple","#0f62fe"), 
  labels = c("RIC", "Death", "Transfer","LTFU"),  
  name = "Outcome event") +
 facet_wrap(~strata, ncol = 1, labeller = labeller(strata = 
      c("ctry=Lesotho" = "Lesotho",
         "ctry=Malawi" = "Malawi", 
        "ctry=Mozambique" = "Mozambique", 
        "ctry=South Africa" = "South Africa",
        "ctry=Zambia" = "Zambia", 
         "ctry=Zimbabwe" = "Zimbabwe"
)))


plot1_dtg0_ctry <- p05 + theme(
         plot.title = element_text(size = 10, hjust = 0.5), #center title
          plot.subtitle = element_text(size = 10, hjust = 0.5), #center subtitle
          axis.title.y = element_text(margin = margin(t = 1, r = 1, b = 1, l = 1), size = 11), #pad y axis label
          axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0), size = 11), #pad x axis label
          axis.text.x = element_text(size = 11, angle=90,vjust = 0.6, hjust = 0.5),
          axis.text.y = element_text(size = 11),
          legend.text = element_text(size = 11),
          legend.title = element_text(size = 11),
          legend.position = "right", 
          legend.justification = "left",
          legend.box.background = element_rect(color = "black"),
          legend.direction="vertical", 
          legend.margin=margin(0,0,0,0),
          legend.box.margin = margin(-5, -5, -5, -5), 
          legend.key.size = unit(0.7, "cm"),
          #strip.background = element_blank(),
          strip.text.x = element_text(size = 12)
          ) +
  guides(color = guide_legend(ncol = 2))  +
    scale_y_continuous(breaks = c(0, 0.25,  0.50, 0.75, 1.0), 
                        labels = c("0%", "25%", "50%", "75%", "100%")) 

plot1_dtg0_ctry

ggsave("output/Abstract_outcomes/Fig1_cohort1_dtg0_outcomes_ctry_calendar_v2.png", width = 5, height =12)
```



<!-- ----------------------------------------------------- -->

# Transition Cohort: Outcomes

```{r eval=FALSE, include=FALSE}

frq(tblBAS_2, dtg_switch)

tblBAS_2[, start_fu := dtg_recomend_d] 
tblBAS_2[, end_fu := rev_outcome_d] 

tblBAS_2[, FU_day := as.numeric(end_fu - start_fu)]
tblBAS_2[, FU_month := round(as.numeric(end_fu - start_fu) / (365.25 / 12), 1)]
tblBAS_2[, FU_year := round(as.numeric(end_fu - start_fu) / (365.25), 1)]

tblBAS_2[, time := FU_day]


```


```{r}
table1::label(tblBAS_2$sex) <- "Sex"
table1::label(tblBAS_2$age_cat3) <- "Age group"

table1::label(tblBAS_2$rev_outcome) <- "Outcome"
table1::label(tblBAS_2$sex_age2) <- "Sex & age group"
table1::label(tblBAS_2$sex_age3) <- "Sex & age group"
table1::label(tblBAS_2$program) <- "Program"
table1::label(tblBAS_2$ctry) <- "Country"

```

```{r}
# create variables for plots showing time as calendar year; all relative to min start date across all cohorts

FU_time_2 <- tblBAS_2 %>% 
  select(patient, program, country, ctry, dtg_recomend_d, dtg, dtg_final, dtg_switch, dtg_switch_txt, rev_outcome, rev_outcome_d, start_fu, end_fu, time) %>%
  mutate(start_min_d = min(start_fu)) %>% # finds min start date across all cohorts (=dtg adoption date)
  mutate(time_t2 = as.numeric(end_fu) - as.numeric(start_min_d)) %>% # gives end time relative to earliest start min date (=earliest dtg adoption date)
  mutate(dtg_recomend_t = as.numeric(dtg_recomend_d) - as.numeric(start_min_d)) %>% # gives dtg adoption date relative to earliest dtg adoption date
  mutate(dtg_comp_b = ifelse(rev_outcome == "RIC", 0, 1))

# create file of dtg_recomend_t
dtg_recomend_t <- FU_time_2 %>% select(program, country, ctry, dtg_recomend_d, dtg_recomend_t) %>% unique()
  
```

```{r include=FALSE}
 # Competing risks regression (alive ==0, dtg ==1, death ==2, ltfu ==3, trans ==4)
  
frq(FU_time_2, rev_outcome)


FU_time_2 <- FU_time_2 %>% 
  mutate(
    
dtg_switch_txt_crr1b = case_when(
    rev_outcome == "RIC" ~ 0, # still alive RIC (retained in care)
    rev_outcome == "Death" ~ 1, # outcome Death
    rev_outcome == "Transfer" ~ 2, # outcome Transfer
    rev_outcome == "LTFU" ~ 3), # outcome LTFU
)

# factor alphabetically, so can determine plot output order of events (_crr1b.factor)
FU_time_2$dtg_switch_txt_crr1b.factor <- factor(FU_time_2$dtg_switch_txt_crr1b, 
levels=c("0", "1", "2", "3"),
labels = c("a", "b", "c", "d"))

```


```{r include=FALSE}
# create two subcohorts: intitated DTG & did not initiate DTG
frq(FU_time_2, dtg)
frq(FU_time_2, rev_outcome)

FU_time_2_DTG <- FU_time_2 %>% filter(dtg_switch == 1)
FU_time_2_noDTG <- FU_time_2 %>% filter(dtg_switch == 0)

```


## All adults 20-39 years

Switched to DTG ART

```{r echo=FALSE}

## by country with DTG
sjPlot::tab_xtab(FU_time_2_DTG$ctry,
                 FU_time_2_DTG$rev_outcome,
                 show.row.prc = TRUE, show.summary = FALSE)

```

Didn't switch to DTG


```{r echo=FALSE}

## by country with no DTG
sjPlot::tab_xtab(FU_time_2_noDTG$ctry,
                 FU_time_2_noDTG$rev_outcome,
                 show.row.prc = TRUE, show.summary = FALSE)
```

<!-- ----------------------------------------------------- -->

# Plots for transition cohort

## Switched to DTG ART by country



```{r}

## Need to fix issue with LSO, ZAF, MOZ not having anyone intiating DTG early...
testLSO <- FU_time_2_DTG %>% filter(country =="LSO") %>% filter(rev_outcome == "RIC")
testMOZ <- FU_time_2_DTG %>% filter(country =="MOZ") %>% filter(rev_outcome == "RIC")

```

```{r}
FU_time_2_DTG <- FU_time_2_DTG %>% mutate(time_t2 = ifelse(patient =="SMLM0004", 124, time_t2))
FU_time_2_DTG <- FU_time_2_DTG %>% mutate(time_t2 = ifelse(patient =="SMMA0042", 305, time_t2))

```


```{r}
 rm(testLSO, testMOZ)
```


Cumulative incidence plot of outcomes among adults switching to DTG ART.


```{r include=FALSE}

fit2_dtg1_ctry <- survfit2(Surv(time_t2, dtg_comp_b) ~ ctry, data = FU_time_2_DTG, etype = dtg_switch_txt_crr1b.factor)

#ggcompetingrisks(fit = fit1_ctry_dtg) + 
 #facet_wrap(~strata, ncol = 1)

ggcompetingrisks(fit = fit2_dtg1_ctry) + 
 facet_wrap(~strata, ncol = 1) +
  geom_vline(data = FU_time_2_DTG, aes(xintercept = dtg_recomend_t),linetype="dashed", size = 0.5) 

#ggcompetingrisks(fit = fit1_ctry_dtg) + 
 #facet_wrap(~strata, ncol = 3)
```


```{r}

p05 <-ggcompetingrisks(fit = fit2_dtg1_ctry,
                 xlab = "", 
                  ylab = "Cumulative incidence (%)",
                 title = "") +
                       #scale_x_continuous(breaks = c(153, 518, 883, 1248, 1613 ), limits = c(0,1613), 
    scale_x_continuous(breaks = c(183, 548, 913, 1278, 1643 ), limits = c(0,1643),  
    labels = c("2019", "2020", "2021", "2022", "2023")) +
  theme_ggsurvfit_default() +
 scale_fill_manual("",
  values = c("#9ef0f0", "gray", "#08bdba",  "purple","#0f62fe"), 
  labels = c("RIC", "Death", "Transfer","LTFU"),  
  name = "Outcome event") +
 facet_wrap(~strata, ncol = 1, labeller = labeller(strata = 
      c("ctry=Lesotho" = "Lesotho",
         "ctry=Malawi" = "Malawi", 
        "ctry=Mozambique" = "Mozambique", 
        "ctry=South Africa" = "South Africa",
        "ctry=Zambia" = "Zambia", 
         "ctry=Zimbabwe" = "Zimbabwe"
)))


plot2_dtg1_ctry <- p05 + theme(
         plot.title = element_text(size = 10, hjust = 0.5), #center title
          plot.subtitle = element_text(size = 10, hjust = 0.5), #center subtitle
          axis.title.y = element_text(margin = margin(t = 1, r = 1, b = 1, l = 1), size = 11), #pad y axis label
          axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0), size = 11), #pad x axis label
          axis.text.x = element_text(size = 11, angle=90,vjust = 0.6, hjust = 0.5),
          axis.text.y = element_text(size = 11),
          legend.text = element_text(size = 11),
          legend.title = element_text(size = 11),
          legend.position = "right", 
          legend.justification = "left",
          legend.box.background = element_rect(color = "black"),
          legend.direction="vertical", 
          legend.margin=margin(0,0,0,0),
          legend.box.margin = margin(-5, -5, -5, -5), 
          legend.key.size = unit(0.7, "cm"),
          #strip.background = element_blank(),
          strip.text.x = element_text(size = 12)
          ) +
  guides(color = guide_legend(ncol = 2))  +
    scale_y_continuous(breaks = c(0, 0.25,  0.50, 0.75, 1.0), 
                        labels = c("0%", "25%", "50%", "75%", "100%")) 
  

plot2_dtg1_ctry

ggsave("output/Abstract_outcomes/Fig1_cohort2_dtg1_outcomes_ctry_calendar.png", width = 5, height =12)
```


## Didn't switch to DTG ART by country


```{r}

## Need to fix issue with LSO, ZAF, MOZ not having anyone intiating DTG early...
testLSO <- FU_time_2_noDTG %>% filter(country =="LSO") %>% filter(rev_outcome == "RIC")


```

```{r}
FU_time_2_noDTG <- FU_time_2_noDTG %>% mutate(time_t2 = ifelse(patient =="SMLM0074", 123, time_t2))

```

```{r}
rm(testLSO)
```


Cumulative incidence plot of outcomes among adults not switching to DTG ART.


```{r include=FALSE}

fit2_dtg0_ctry <- survfit2(Surv(time_t2, dtg_comp_b) ~ ctry, data = FU_time_2_noDTG, etype = dtg_switch_txt_crr1b.factor)

#ggcompetingrisks(fit = fit1_ctry_dtg) + 
 #facet_wrap(~strata, ncol = 1)

ggcompetingrisks(fit = fit2_dtg0_ctry) + 
 facet_wrap(~strata, ncol = 1) +
  geom_vline(data = FU_time_2_noDTG, aes(xintercept = dtg_recomend_t),linetype="dashed", size = 0.5) 

#ggcompetingrisks(fit = fit1_ctry_dtg) + 
 #facet_wrap(~strata, ncol = 3)
```


```{r}

p05 <-ggcompetingrisks(fit = fit2_dtg0_ctry,
                 xlab = "", 
                  ylab = "Cumulative incidence (%)",
                 title = "") +
                  #scale_x_continuous(breaks = c(153, 518, 883, 1248, 1613 ), limits = c(0,1613), 
    scale_x_continuous(breaks = c(183, 548, 913, 1278, 1643 ), limits = c(0,1643), 
    labels = c("2019", "2020", "2021", "2022", "2023")) +
  theme_ggsurvfit_default() +
 scale_fill_manual("",
  values = c("#9ef0f0", "gray", "#08bdba",  "purple","#0f62fe"), 
  labels = c("RIC", "Death", "Transfer","LTFU"),  
  name = "Outcome event") +
 facet_wrap(~strata, ncol = 1, labeller = labeller(strata = 
      c("ctry=Lesotho" = "Lesotho",
         "ctry=Malawi" = "Malawi", 
        "ctry=Mozambique" = "Mozambique", 
        "ctry=South Africa" = "South Africa",
        "ctry=Zambia" = "Zambia", 
         "ctry=Zimbabwe" = "Zimbabwe"
)))


plot2_dtg0_ctry <- p05 + theme(
         plot.title = element_text(size = 10, hjust = 0.5), #center title
          plot.subtitle = element_text(size = 10, hjust = 0.5), #center subtitle
          axis.title.y = element_text(margin = margin(t = 1, r = 1, b = 1, l = 1), size = 11), #pad y axis label
          axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0), size = 11), #pad x axis label
          axis.text.x = element_text(size = 11, angle=90,vjust = 0.6, hjust = 0.5),
          axis.text.y = element_text(size = 11),
          legend.text = element_text(size = 11),
          legend.title = element_text(size = 11),
          legend.position = "right", 
          legend.justification = "left",
          legend.box.background = element_rect(color = "black"),
          legend.direction="vertical", 
          legend.margin=margin(0,0,0,0),
          legend.box.margin = margin(-5, -5, -5, -5), 
          legend.key.size = unit(0.7, "cm"),
          #strip.background = element_blank(),
          strip.text.x = element_text(size = 12)
          ) +
  guides(color = guide_legend(ncol = 2))  +
    scale_y_continuous(breaks = c(0, 0.25,  0.50, 0.75, 1.0), 
                        labels = c("0%", "25%", "50%", "75%", "100%")) 

plot2_dtg0_ctry

ggsave("output/Abstract_outcomes/Fig1_cohort2_dtg0_outcomes_ctry_calendar.png", width = 5, height =12)
```


<!-- ----------------------------------------------------- -->

# Combine plots

## Initiation cohort

### Using patchwork (with title)


```{r echo=FALSE}


FigA <- (plot1_dtg1_ctry + plot1_dtg0_ctry + plot_layout(ncol = 2, width =c(9,9))) + plot_layout(guides = "collect") + plot_annotation(title = '(A) Initiating dolutegravir        (B) Not initiating dolutegravir', 
                  caption = '',
                  theme=theme(plot.title=element_text(hjust=0, size = 14))) & theme(legend.position = "right", legend.justification = "center")

FigA
  
  

#plot1_dtg1_ctry

#plot1_dtg0_ctry

ggsave("output/Abstract_outcomes/A_Fig1_cohort1_ctry_sex20-39_calendar.png", width = 6, height =8)

```


## Transition cohort

### Using patchwork (with title)


```{r echo=FALSE}


FigB <- (plot2_dtg1_ctry + plot2_dtg0_ctry + plot_layout(ncol = 2, width =c(9,9))) + plot_layout(guides = "collect") + plot_annotation(title = '(A) Switching to dolutegravir   (B) Not switching to dolutegravir', 
                  caption = '',
                  theme=theme(plot.title=element_text(hjust=0, size = 14))) & theme(legend.position = "right", legend.justification = "center")

FigB
  
  

#plot1_dtg1_ctry

#plot1_dtg0_ctry

ggsave("output/Abstract_outcomes/B_Fig1_cohort2_ctry_sex20-39_calendar.png", width = 6, height =8)

```


<!-- ----------------------------------------------------- -->

# Computing Environment

```{r echo=FALSE, results='asis'}
report(sessionInfo())
```